---
title: "Creating Composite Scores from Multi-Item Measures"
subtitle: "How to create a composite and check reliability of the measure"
author: Shane McCarty, PhD
abstract: |
  This chapter teaches researchers how to create composite scores from multi-item 
  scales and assess their reliability. Using the Regulatory Focus Questionnaire 
  (Higgins, 2001) as an example, researchers learn to measure promotion focus 
  (preference for growth) and prevention focus (preference for maintaining status quo) 
  through a systematic four-step process: (1) define which survey items belong to 
  each construct, (2) identify and reverse-code negatively worded items, (3) calculate 
  composite scores by averaging items using the psych package's scoreItems() function, 
  and (4) assess scale reliability using Cronbach's alpha. The chapter explains why 
  single items are insufficient for measuring psychological constructs and demonstrates 
  how multiple items averaged together create more reliable continuous measures. 
  Researchers learn to interpret Cronbach's alpha values (with acceptable reliability 
  typically above .70) and understand when reverse coding is necessary for items that 
  capture the absence rather than presence of a construct. Through practical examples 
  with health beliefs data, researchers master creating composite scores that accurately 
  reflect theoretical constructs while handling missing data and ensuring all 
  participants' scores have equivalent meaning. This skill is essential when working 
  with validated multi-item scales in public health and behavioral research.
keywords:
  - composite scores
  - Cronbach's alpha
  - reliability
  - survey scales
format: html
editor: visual
execute:
  warning: false
  message: false
---

::: callout-tip
## 📖 `psych` resources

-   [Documentation](https://cran.r-project.org/web/packages/psych/psych.pdf)
-   [Personality Project in R](https://personality-project.org/r/psych/)
-   [Reliability](https://rpubs.com/deondb/reliability)
-   [Reliability Analysis](https://rpubs.com/hauselin/reliabilityanalysis)
:::

## Dataset

The health beliefs dataset includes a range of variables, including promotion focus and prevention focus.

-   Promotion focus: a general regulatory motivational preference toward growth and change, attention to positive things, and success seeking

-   Prevention focus: a general regulatory motivational preference toward maintaining the status quo, avoiding negative things, and aversion to threats

### Load Library

```{r}
#| label: load-library
library(dplyr)
library(psych)
library(knitr)
library(ggplot2)
library(readxl)
```

### Import

```{r}
#| label: import-excel-header1
#| autorun: true
#| completion: true
#| echo: true
#| output: false
#| warning: false
#| error: false

# Import Excel file
healthbeliefsdata <- read_excel("data/03.15.2025.prolifichealthbeliefs.xlsx",
    col_names = TRUE)

healthbeliefsdata[healthbeliefsdata == -99] <- NA
healthbeliefsdata[healthbeliefsdata == -50] <- NA

##explanation: all -99 and -50 data will be treated as missing data

```

### Scoring Key to Create Composites

In psychology, researchers aim to measure a construct using multiple items. For example, promotion focus (`PROMOFOCUS`) and prevention focus (`PREVFOCUS`) are two separate measures with 6 items and 5 items, respectively. To create a composite measure, each item is added and divided by the number of items, resulting in an average score for the measure.

-   `PROMOFOCUS`: ( `FOCUS_1` + `FOCUS_2` + `FOCUS_3` + `FOCUS_4` + `FOCUS_5` + `FOCUS_6` ) / 6
-   `PREVFOCUS`: ( `FOCUS_7` + `FOCUS_8` + `FOCUS_9` + `FOCUS_10` + `FOCUS_11` ) / 5

No single item is perfectly predictive of a construct. In rare cases, a single item can approximate a construct, such as the single item self-esteem scale (Robbins et al., 2001) for self esteem and the single item life satisfaction measure. In most cases, a single item is highly problematic. For this reason (and to create a continuous -- not ordinal), multiple items are averaged to create a **composite score**.

```{r}
#| label: create-keys

library(psych)
# Create keys for scoring
focus_keys <- list(
  PROMOFOCUS = c("FOCUS_1", "FOCUS_2", "FOCUS_3", "FOCUS_4", "FOCUS_5", "FOCUS_6"),
  PREVFOCUS = c("FOCUS_7", "FOCUS_8", "FOCUS_9", "FOCUS_10", "FOCUS_11") 
)

#explanation: this creates a list called focus_keys which tells R which survey questions belong to which composite score (promotion or prevention)
```

::: callout-note
## What is focus_keys?

-   Items 1-6 measure promotion focus (`PROMOFOCUS)`
-   Items 7-11 measure prevention focus (`PREVFOCUS)`
:::

### Scoring Key with Reverse Items

#### Affirmative Items

Below are two examples of FOCUS items:

-   Promotion Item (`FOCUS_3`): "How often have you accomplished things that got you "psyched" to work even harder?"

-   Prevention Item (`FOCUS_9`): "How often did you obey rules and regulations that were established by your parents?"

These items were measured on a 5-point scale:

-   Never or Seldom (1)
-   Rarely (2)
-   Sometimes (3)
-   Often (4)
-   Very often (5)

#### Negative Items (aka Reverse-Coded Items)

-   Promotion Item (`FOCUS_6`): "I have found very few hobbies or activities in my life that capture my interest or motivate me to put effort into them."
-   Prevention Item (`FOCUS_8`): "Did you get on your parents’ nerves often when you were growing up?"

Most often, a construct is measured with **affirmative items** where the most agreement/occurrence (very often) is coded as a 5 on a 1-5 scale. Sometimes, a construct also includes negative items where the strongest agreement (very often) to a **negative item** is coded as 1 on a 1-5 scale. In essence, for reverse-scored items, the coded values are flipped.

Notice how the phrase "**very few"** in the item -- "I have found **very few** hobbies or activities in my life that capture my interest or motivate me to put effort into them." -- captures the *absence* of promotion focus.

On the original scale, "Very often" = 5. However, because this is a **negatively worded item**, we reverse the scoring:

-   Original 5 → Reversed to 1

-   Original 4 → Reversed to 2

-   Original 3 → Stays 3

-   Original 2 → Reversed to 4

-   Original 1 → Reversed to 5

Therefore, if someone reports "Very often" (5) for `FOCUS_6`, after reverse coding it becomes 1, correctly reflecting LOW promotion focus.

::: callout-important
## Understanding the `-` symbol

The minus sign (`-`) before an item name tells `scoreItems()` to reverse code that item.

Example: - `"FOCUS_6"` = use original scores - `"-FOCUS_6"` = reverse the scores (1→5, 2→4, 3→3, 4→2, 5→1)
:::

```{r}
#| label: reversecoding
# reverse coding items based on scoring key from here: https://sjdm.org/dmidi/Regulatory_Focus_Questionnaire.html . Our survey did NOT put the items in the same order which is why it is different than the downloadable word doc.
focus_keys_with_reverse <- list(
  PROMOFOCUS = c("-FOCUS_1", "FOCUS_2", "FOCUS_3", "-FOCUS_4", "FOCUS_5", "-FOCUS_6"),
  PREVFOCUS = c("-FOCUS_7", "-FOCUS_8", "FOCUS_9", "-FOCUS_10", "-FOCUS_11") # this part includes reverse-coded items
)
```

```{r}
#| label: create-composite-withreversed

library(psych)
# Using scoreItems() - recommended
focus_scores <- scoreItems(focus_keys_with_reverse, healthbeliefsdata) 

# Extract the scores
composite_scores <- focus_scores$scores

# Add to your dataframe
healthbeliefsdata$PROMOFOCUS <- composite_scores[, "PROMOFOCUS"]
healthbeliefsdata$PREVFOCUS <- composite_scores[, "PREVFOCUS"]
```

## Reliability Analysis

A reliability analysis of the **composite measure(s)** determines if the items accurately fact reflect your theoretical/conceptual construct. The indicator ranges from 0 to 1 with 1 reflecting higher correlations among the items.

```{r}
#| label: reliability analysis
# View reliability statistics
focus_scores
```

### Interpreting Cronbach's Alpha

| Alpha Value   | Interpretation                                    |
|---------------|---------------------------------------------------|
| α \< 0.60     | Unacceptable - items don't measure the same thing |
| α = 0.60-0.69 | Questionable - marginal reliability               |
| α = 0.70-0.79 | Acceptable - adequate reliability                 |
| α = 0.80-0.89 | Good - strong reliability                         |
| α ≥ 0.90      | Excellent - very high reliability                 |

**Our results:**

-   Promotion Focus (α = .68): **Questionable** - barely acceptable

-   Prevention Focus (α = .74): **Acceptable** - adequate reliability

::: callout-tip
## What if alpha is too low?

-   Check for reverse-coded items you missed
-   Look for items that don't fit conceptually
-   Consider removing problematic items
-   Use at least 3-4 items per scale
:::

::: callout-important
## Requirement

In the measures section of your `RDreport.qmd` and final report, please include the results of the reliability analysis (aka cronbach's alpha).
:::

::: callout-warning
This page is under construction. It will be updated in the future!
:::
