---
title: "Describe Your Data"
subtitle: "How to check whether your outcome is normally distributed"
author:
  - Shane McCarty, PhD
abstract: |
  This chapter is the first step on the Analysis Map. Researchers check the distribution of the outcome in their research question with a histogram, a Q-Q plot, and the Shapiro-Wilk test, and then decide whether to use a parametric test (normal data) or a non-parametric test (data that are not normal).
keywords:
  - normality
  - distribution
  - histogram
  - Q-Q plot
  - Shapiro-Wilk
  - parametric
  - non-parametric
date: 09.20.2026
date-format: "MM.DD.YYYY"
format:
  html:
    css: custom.css
editor: visual
execute:
  cache: false   # this chapter reads lab_prep.R and the lab dataset; a cache would keep OLD data after they change
  warning: false
  message: false
---

{{< include _ref-kickoff-box.qmd >}}

## Why check for normality?

Most statistical tests come in two versions.

-   **Parametric tests** (t-test, ANOVA, Pearson correlation) assume that your outcome is **normally distributed**. On a histogram, normal data look like a bell: most people are in the middle, and fewer people are at each end.
-   **Non-parametric tests** (Mann-Whitney U, Wilcoxon, Kruskal-Wallis, Spearman correlation) do not make that assumption.

So before you choose a test on the [Analysis Map](analysis-map.qmd), you need one answer: **is my outcome normal, or not?**

::: {.callout-tip}
## Go: Check the outcome in your research question

You do not need to check every variable in your dataset. Check the **outcome**: the score your research question is about.
:::

## The example data

This chapter uses the lab dataset and three outcomes:

-   `WELLBEING`: the average of eight items (a composite score from 1 to 5)
-   `STIGMA`: the average of three items (a composite score from 1 to 5)
-   `HELPSEEK`: one item with answers from 1 to 5

```{r}
#| label: load-normality
library(tidyverse)
source("lab_prep.R")   # creates mh_clean
```

## Check 1: Look at a histogram

Always start here. Ask: is it roughly bell-shaped, with one hump in the middle?

```{r}
#| label: plot-hist-wellbeing
#| fig-cap: "Figure 1. Histogram of the well-being composite score."
#| fig-alt: "A histogram of well-being scores with one hump in the middle and a similar tail on each side."
#| fig-height: 4
#| fig-width: 6
ggplot(data = mh_clean,
       mapping = aes(x = WELLBEING)) +
  geom_histogram(binwidth = 0.25, fill = "#005a43", color = "white") +
  xlab("Positive mental well-being (1 to 5)") +
  ylab("Number of people") +
  theme_bw()
```

```{r}
#| label: plot-hist-stigma
#| fig-cap: "Figure 2. Histogram of the stigma composite score."
#| fig-alt: "A histogram of stigma scores with one hump below the middle of the scale and a longer tail toward the high end."
#| fig-height: 4
#| fig-width: 6
ggplot(data = mh_clean,
       mapping = aes(x = STIGMA)) +
  geom_histogram(binwidth = 1/3, fill = "#005a43", color = "white") +
  xlab("Stigma (1 to 5)") +
  ylab("Number of people") +
  theme_bw()
```

```{r}
#| label: plot-hist-helpseek
#| fig-cap: "Figure 3. Histogram of the single help-seeking item."
#| fig-alt: "A bar-like histogram of a single item with five possible answers, piled up toward one end of the scale."
#| fig-height: 4
#| fig-width: 6
ggplot(data = mh_clean,
       mapping = aes(x = HELPSEEK)) +
  geom_histogram(bins = 5, fill = "#005a43", color = "white") +
  xlab("Help seeking (1 to 5)") +
  ylab("Number of people") +
  theme_bw()
```

::: {.callout-warning}
## Caution: A single survey item is not normal

One item with only five possible answers cannot make a smooth bell. Treat a single item as **not normal** and use the non-parametric test. A composite score (the average of several items) has many more possible values and is often close to normal. The more items in the composite, the smoother the histogram.
:::

## Check 2: Look at a Q-Q plot

A Q-Q plot compares your data to a perfect normal distribution. If the points follow the line, the data are close to normal. If the points bend away from the line, especially at the ends, the data are not normal.

```{r}
#| label: plot-qq-wellbeing
#| fig-cap: "Figure 4. Q-Q plot of the well-being composite score."
#| fig-alt: "A Q-Q plot where the points stay close to the diagonal line from end to end."
#| fig-height: 4
#| fig-width: 5
ggplot(data = mh_clean,
       mapping = aes(sample = WELLBEING)) +
  geom_qq() +
  geom_qq_line() +
  theme_bw()
```

```{r}
#| label: plot-qq-stigma
#| fig-cap: "Figure 5. Q-Q plot of the stigma composite score."
#| fig-alt: "A Q-Q plot where most points follow the diagonal line and the points at the ends bend away from it."
#| fig-height: 4
#| fig-width: 5
ggplot(data = mh_clean,
       mapping = aes(sample = STIGMA)) +
  geom_qq() +
  geom_qq_line() +
  theme_bw()
```

## Check 3: Run the Shapiro-Wilk test

The Shapiro-Wilk test puts a number on what you saw in the plots.

```{r}
#| label: shapiro-wellbeing
shapiro.test(mh_clean$WELLBEING)
```

```{r}
#| label: shapiro-stigma
shapiro.test(mh_clean$STIGMA)
```

```{r}
#| label: shapiro-helpseek
shapiro.test(mh_clean$HELPSEEK)
```

How to read the p-value. This one feels backwards, so read it slowly.

-   **p is larger than .05** → the data are **not** different from normal → treat the outcome as **normal**.
-   **p is smaller than .05** → the data **are** different from normal → treat the outcome as **not normal**.

**Reading the example.** `WELLBEING` is **normal**: the histogram is a bell, the Q-Q points follow the line, and the Shapiro-Wilk p-value (`r sprintf("%.3f", shapiro.test(mh_clean$WELLBEING)$p.value)`) is larger than .05. `HELPSEEK` is also easy: the histogram is piled up at one end, and p is far below .05. It is not normal. `STIGMA` is harder. The histogram has one hump, but the tail on the right is longer, and the Shapiro-Wilk p-value (`r sprintf("%.3f", shapiro.test(mh_clean$STIGMA)$p.value)`) is below .05. The checks do not fully agree, so the safe decision is **not normal**.

::: {.callout-warning}
## Caution: Do not use the Shapiro-Wilk test by itself

With a large sample, the test flags tiny differences that do not matter. With a small sample, it can miss big ones. Use the histogram and the Q-Q plot together with the test. When the three checks disagree, or when you are unsure, choose the non-parametric test and talk to your instructor.
:::

## Make the decision

| What you found | Your outcome is | Use the |
|------------------------------|------------------|--------------------|
| Bell-shaped histogram, points follow the Q-Q line, Shapiro-Wilk p larger than .05 | normal | **parametric** test (listed under NORMAL on the Analysis Map) |
| Skewed or lumpy histogram, points bend away from the Q-Q line, Shapiro-Wilk p smaller than .05 | not normal | **non-parametric** test (listed under NOT NORMAL on the Analysis Map) |
| A single survey item (for example, 1 to 5) | not normal | **non-parametric** test |

Write your answer in your data analysis plan, and report it in your methods:

> The distribution of \_\_\_\_\_\_\_\_ was checked with a histogram, a Q-Q plot, and the Shapiro-Wilk test (W = \_\_\_, p = \_\_\_). The data were \_\_\_\_\_\_\_\_ (normal / not normal), so a \_\_\_\_\_\_\_\_ test was used.

::: {.callout-warning}
## Caution: Don't pick the test that gives you the result you want

Decide normal or not normal **before** you run your statistical test, and keep that decision.
:::

::: callout-note
## One exception: regression

For a linear regression, you check the **residuals** of the model, not the raw outcome. See [Relate 3+ Variables](regression.qmd).
:::

{{< include _ref-whistle-box.qmd >}}
