---
title: "Transforming Your Data"
subtitle: "How to use dplyr"
author: "Mara Estreich"
abstract: |
  This chapter introduces students to dplyr, a core package within the tidyverse designed
  for efficient data manipulation and transformation. Students learn essential data cleaning
  techniques including filtering observations, recoding variables, and creating new categorical
  and binary variables. Through practical examples using survey data, the chapter demonstrates
  how to prepare datasets for analysis by grouping response categories and creating dummy
  variables for regression modeling. By the end of this chapter, students will understand
  that clean, well-organized data is foundational for accurate analysis and will be equipped
  with core dplyr functions to transform raw survey data into analysis-ready datasets.
keywords:
  - data manipulation
  - dplyr
  - recoding variables
  - creating new variables
  - dummy coding
format: html
editor: visual
execute:
  warning: false
  message: false
---

::: callout-tip
## 📖 Learning Resources

-   **R for Data Science**: [Data Transformation](https://r4ds.hadley.nz/data-transform.html)
-   **R for Data Science**: [Transform](https://r4ds.hadley.nz/transform.html)
-   **Datacamp Course**: [dplyr](https://www.datacamp.com/courses/data-manipulation-with-dplyr)
-   **Cheat Sheet**: [dplyr](https://rstudio.github.io/cheatsheets/data-transformation.pdf)
-   **Posit Interactive Page**: [dplyr](https://rstudio.github.io/cheatsheets/html/data-transformation.html)
:::

```{r}
#| label: load library
library(haven)
library(naniar)
library(dplyr)
library(knitr)
```

```{r}
#| label: import-dataset
alldata <- read_sav("data/10.31.2024.oralhealth.sav", user_na = FALSE)
```



```{r}
#| label: replace-missing-data

## replace with NA
alldata <- alldata %>%
  replace_with_na_all(condition = ~.x %in% c(-99, -50))
```

## Introduction to dplyr

dplyr is part of the tidyverse and provides intuitive functions for manipulating data. Think of it as a toolkit for cleaning and preparing your data for analysis.

**Why use dplyr?**

-   Makes data cleaning code readable and logical

-   Uses the pipe operator (`%>%`) to chain operations together

-   Essential for preparing data before creating visualizations or running analyses

> **Remember**: You cannot create meaningful visualizations or run valid statistical analyses without clean data!

## Example 1: Recoding Education into Categories

When working with survey data, we often need to collapse detailed response categories into broader groups. This is useful for:

-   Simplifying analysis

-   Creating groups with adequate sample sizes

-   Making results easier to interpret

### Original Education Variable

Our survey asked about education level with 7 response options:

1.  Less than high school

2.  High school diploma or GED

3.  Some college, but no degree

4.  Associate's or technical degree

5.  Bachelor's degree

6.  Master's degree

7.  Doctoral or professional degree

### Recoding into Three Categories

Let's group these into three meaningful categories:

```{r}
#| label: mutate-education-to-education_3cat
  
# Create a new 3-category education variable
alldata <- alldata %>%
  mutate(
    education_3cat = factor(
      case_when(
        Education_Level %in% c(1, 2, 3, 4) ~ "Less than Bachelor's",
        Education_Level == 5 ~ "Bachelor's Degree",
        Education_Level %in% c(6, 7) ~ "Graduate Degree"
      ),
      levels = c("Less than Bachelor's", "Bachelor's Degree", "Graduate Degree")
    )
  )

# Check the recoding with a cross-tabulation
table(alldata$Education_Level, alldata$education_3cat, useNA = "ifany")
```

::: callout-warning
This page is under construction. It will include more information soon!
:::

