---
title: "Import Data Once"
subtitle: "How to use readr, readxl, and haven"
authors: 
  - Zihan Hei
  - Shane McCarty, PhD
abstract: |
  This chapter supports researchers in importing data from various sources and file formats commonly used in public health research. Researchers learn to load datasets from R packages (NHANES), import Excel spreadsheets, read csv files, and work with SPSS data files. The chapter demonstrates essential data inspection techniques including viewing data structure, examining headers, and understanding data summaries. Researchers also learn to handle missing data by recoding special values (-99, -50) as NA during import. By the end of this chapter, researchers will be equipped to import data from multiple sources and prepare it for analysis in their public health research projects.
keywords:
  - data import
  - csv
  - excel
  - SPSS
  - NHANES
  - missing data
date: 10.20.2025
date-format: "MM.DD.YYYY"
format:
  live-html:
    css: custom.css
engine: knitr
editor: visual
execute:
  cache: false   # reads the lab dataset; also, an old cache listed packages (NHANES) that are not installed on every computer
---

{{< include ./_extensions/r-wasm/live/_knitr.qmd >}}

::: {.callout-note collapse="true"}
## Resources: import

- [R4DS Book Chapter 4](https://r4ds.hadley.nz/data-import.html)
- [R4DS Book Chapter 20-24](https://r4ds.hadley.nz/import.html)
- [Posit Cheatsheet](https://rstudio.github.io/cheatsheets/html/data-import.html?_gl=1*2grk0d*_ga*ODQ0NjUzMTM1LjE3NjA0MDU5Nzk.*_ga_2C0WZ1JHG0*czE3NjA0OTMzNDgkbzMkZzAkdDE3NjA0OTMzNDgkajYwJGwwJGgw)
- [PDF Cheatsheet](https://rstudio.github.io/cheatsheets/data-import.pdf)
:::


```{r}
#| include: false
current_date <- format(Sys.Date(), "%m.%d.%Y")

```

## Introduction to Data Import

The first step of the data science workflow is to *import your data*. Public health researchers work with data from many sources and in many formats. In FRI-PH, we will import data from a survey collected using the survey software Qualtrics.

**Common data sources:**

- Survey platforms (Qualtrics, REDCap)

- R packages (e.g., `NHANES`)

- Other statistical softwares (e.g., SPSS, SAS, Stata)

- Spreadsheets (Excel, Google Sheets)

- Databases (SQL)

```{r}
#| label: load-packages
#| eval: false
#| autorun: true
#| completion: true
#| warning: false
#| message: false

# Load packages (do this each session)
library(NHANES)    # CDC health data
library(tidyverse) # Data wrangling (includes readr)
library(readxl)    # Read Excel files
library(haven)     # Read SPSS/SAS/Stata files
```

## Datasets

You will learn how to import different datasets:

- datasets in R packages `NHANES` and `YRBS`

- class dataset "health beliefs (cohort 11) collected in the Research Methods I (ANTH206) course

- team dataset "oralhealth"

### File Formats

- National Health and Nutrition Examination Survey dataset from the Centers for Disease Control and Prevention (CDC) using the R package `NHANES`

- a `.xlsx` file format of the class dataset "health beliefs (cohort 11) collected in the Research Methods I (ANTH206) course `03.15.2025.prolifichealthbeliefs.xlsx`

- a `.csv` file format of the class dataset "health beliefs (cohort 11)" collected in the Research Methods I (ANTH206) course `03.15.2025.prolifichealthbeliefs.csv`

- a `.sav` file format from SPSS of the FRI team project on oral health `10.31.2024.oralhealth.sav`.

## Import Examples

### Import from `NHANES` package

```{r}
#| label: install-NHANES
#| eval: false
install.packages("NHANES")
```

### Load libraries

```{r}
#| label: load-libraries-example
#| eval: false
#| autorun: false
#| completion: true
#| echo: true
#| output: true
#| warning: false
#| error: false

library(tidyverse)
library(psych)
library(knitr)
library(tibble)
library(dplyr)
library(tidyr)
library(scales)     # for number formatting like comma()
library(english)    # to convert numbers to words
library(stringr)    # for text functions like str_c()
library(NHANES)
```

#### Column Names

Review the column names (aka the variables) in the dataset.

```{webr-r}
#| label: names-NHANES
library(NHANES)
names(NHANESraw)
```

#### Data Structure

Now, you will view the structure of the NHANES data

```{webr-r}
#| label: data-structure-NHANES
str(NHANESraw)
```

#### Data Header

Review the first 10 responses for the first 6 columns

```{webr-r}
library(NHANES)
NHANESraw[1:10, 1:6] 
```

#### View Data

In RStudio, you can use `View()` in your CONSOLE to view the dataframe.

```{{r}}
library(NHANES)
# View data
View(NHANESraw)
```

### Import Excel `.xlsx` file {#row1}

Use the `readxl` package to import Excel files (.xlsx or .xls). The primary advantage of using Excel files is the ability to manually inspect your data using Excel or Google sheets before you import into R. Below are code chunks that can be copied into your report if you are using an excel file with the **header in row 1**.

::: callout-important
## Important: Use the `.clean` data file

Use the .clean data file: `r current_date`.data.team1.clean.xlsx with a header in row 1 only. If you manually inspect your data and you see labels in row 2, you need to clean the dataset.
:::

```{r}
#| label: import-excel-header1
#| autorun: true
#| completion: true
#| echo: true
#| output: false
#| warning: false
#| error: false

library(readxl)
# Import Excel file
safetydata <- read_excel(
    "10.20.2025.data.team1.clean.xlsx",
    col_names = TRUE)

safetydata[safetydata == -99] <- NA
safetydata[safetydata == -50] <- NA

##explanation: all -99 and -50 data will be treated as missing data
```

```{r}
# View first 10 rows
head(safetydata, 10)
```

```{r}
# View column names, which are the VARIABLES
names(safetydata)
```

```{r}
# Get a summary of all variables
summary(safetydata)
```

```{r}
#| label: lab-import-setup
#| include: false
#| cache: false
library(readxl)
mh_raw_check <- read_excel("data/ANTH306_LayConceptionsMH_SYNTHETIC.xlsx", sheet = "data")
mh_codebook_check <- read_excel("data/ANTH306_LayConceptionsMH_SYNTHETIC.xlsx", sheet = "codebook")
```

## 🏈 The Lab {#sec-import-lab}

You watched the play above with example datasets. Now import the **lab dataset** yourself: a survey about what people believe about mental health. Everyone in the course uses the same file, so you can check your answers.

:::::: {.ref .ref-tip}
::: ref-face
![](images/ref.svg){fig-alt="Ref the raccoon"}
:::

::: ref-say
**Ref's kickoff.** This lab is not graded, so I am here to help you check your own work. Try each step **before** you open my check. If your answer does not match, that is not a penalty. It is how you find out what to fix.
:::
::::::

**Before you start**

1.  Get the lab data file, `ANTH306_LayConceptionsMH_SYNTHETIC.xlsx`, from your instructor. Put it in a folder called **data** inside your `Lastname_FRI` project folder.
2.  Open your RStudio Project and your lab `.qmd` file, and run your `library()` chunk. You need `readxl`.

### Lab Play 1: Import the data

The file has more than one sheet. The survey answers are on the sheet called `data`. Import that sheet and name the object `mh_raw`. Then find out how many rows (people) and columns (variables) it has.

::: {.callout-tip collapse="true" icon="false"}
## ![](images/ref.svg){.ref-mini fig-alt="Ref the raccoon"} Ref's check: how big is the dataset?

`mh_raw` should have **`r nrow(mh_raw_check)` rows** and **`r ncol(mh_raw_check)` columns**. Look in your **environment tab**: it shows the same two numbers next to `mh_raw`.
:::

::: {.callout-note .gray collapse="true"}
## 👀 See the solution

```{r}
#| label: lab-import-solution
#| eval: false
library(readxl)

mh_raw <- read_excel("data/ANTH306_LayConceptionsMH_SYNTHETIC.xlsx",
                     sheet = "data")

nrow(mh_raw)   # how many people?
ncol(mh_raw)   # how many variables?
```
:::

### Lab Play 2: View the variables

Look at the column names with `names()`, look at the first rows with `head()`, and get a summary of one variable, `POLITICALBELIEFS`, with `summary(mh_raw$POLITICALBELIEFS)`.

::: {.callout-tip collapse="true" icon="false"}
## ![](images/ref.svg){.ref-mini fig-alt="Ref the raccoon"} Ref's check: what should you see?

-   The first five column names are: `r paste0("\x60", names(mh_raw_check)[1:5], "\x60", collapse = ", ")`. The last one is `r paste0("\x60", tail(names(mh_raw_check), 1), "\x60")`.
-   `POLITICALBELIEFS` has a minimum of `r min(mh_raw_check$POLITICALBELIEFS, na.rm = TRUE)` and a maximum of `r max(mh_raw_check$POLITICALBELIEFS, na.rm = TRUE)`.
-   A minimum of -99 is not a real answer. The answers to this question run from 1 to 6, and -99 is the code for "Prefer not to say". You will turn those codes into missing values (`NA`) the same way this chapter did for `safetydata`.
:::

### Lab Play 3: Do the variable names match the survey?

This is the check that matters most. The column names in R come straight from the variable names that were typed into Qualtrics. If a name is misspelled, or a question was never named, you will see it here first (for example, a column called `Q17` and no column called `STIG3`).

For the lab dataset, the survey's variable names are listed on the sheet called `codebook`. Import that sheet, and compare its `Variable` column with `names(mh_raw)`.

::: {.callout-tip collapse="true" icon="false"}
## ![](images/ref.svg){.ref-mini fig-alt="Ref the raccoon"} Ref's check: do they match?

```{r}
#| label: lab-import-check-names
#| echo: false
#| cache: false
cb_names <- mh_codebook_check$Variable
missing_from_data <- setdiff(cb_names, names(mh_raw_check))
```

-   The survey variables in `mh_raw` should appear in the codebook, spelled the same way.
-   `setdiff(names(mh_raw), mh_codebook$Variable)` should return **`r length(setdiff(names(mh_raw_check), cb_names))` names**, and all of them should start with `r paste0("\x60", unique(sub("_[0-9]+$", "", setdiff(names(mh_raw_check), cb_names))), "\x60", collapse = ", ")`. The codebook describes that group of columns in one line, so they do not match one by one. That is fine.
-   If you see any **other** name in that list, the data and the survey do not agree, and you would need to find out why.
:::

::: {.callout-note .gray collapse="true"}
## 👀 See the solution

```{r}
#| label: lab-import-solution-names
#| eval: false
names(mh_raw)                  # the VARIABLES in your dataset
head(mh_raw, 10)               # the first 10 people
summary(mh_raw$POLITICALBELIEFS)   # one variable: look at the minimum

mh_codebook <- read_excel("data/ANTH306_LayConceptionsMH_SYNTHETIC.xlsx",
                          sheet = "codebook")

# names that are in the data and NOT in the codebook
setdiff(names(mh_raw), mh_codebook$Variable)
```
:::

::: {.callout-warning collapse="true" icon="false"}
## ![](images/ref-whistle.svg){.ref-mini fig-alt="Ref the raccoon blowing his whistle"} Ref's call: fouls to check if your numbers do not match

-   **`path does not exist`** → the file is not where the code says it is. Check that the file is inside a **data** folder in your project, that the name is spelled exactly the same, and that you opened your **project** (not only a file).
-   **You have 1 column, or the wrong columns** → you imported the wrong sheet. Add `sheet = "data"`.
-   **`could not find function "read_excel"`** → run `library(readxl)`.
-   **Your first row of data is question text** → the file still has two header rows. See [Export Survey Data](export.qmd) and delete row 2 in the `.clean` file. (The lab dataset has one header row, so you should not see this in the lab.)
:::

:::::: {.ref .ref-tip}
::: ref-face
![](images/ref.svg){fig-alt="Ref the raccoon"}
:::

::: ref-say
**Ref's final whistle.** That is the end of the lab. Save your `.qmd`, render it, and back up your `.qmd` and `.R` files to the **R Labs** folder in your ELN before you quit RStudio.
:::
::::::

## 🏆 Your Turn {#sec-import-yourturn}

::: {.callout-note .gray icon="false"}
## There is no Ref. It's game time, your turn!

Nobody has imported your team's data before, so there is no answer to check against. Use the checklist below.
:::

Once you finish all of the R Labs, come back here with **your team's `.clean` data file**.

```{r}
#| label: import-REPLACE
#| eval: false
library(readxl)

yourdata <- read_excel("data/REPLACE.data.team#.clean.xlsx")   # REPLACE with your file name

yourdata[yourdata == -99] <- NA    # prefer not to answer -> missing
yourdata[yourdata == -50] <- NA    # don't know -> missing

names(yourdata)      # the VARIABLES
head(yourdata, 10)   # the first 10 people
summary(yourdata)    # minimum and maximum of every variable
```

| Criteria | Ask yourself |
|------------|--------------------------------------------------------|
| **One header row** | Is row 1 the variable names, and is row 2 a person (not question text)? |
| **Rows** | Does the number of rows match the number of responses in Qualtrics? |
| **Variable names** | Open your exported survey (`.docx`) next to `names(yourdata)`. Does every column name match the variable name you chose in Qualtrics, spelled the same way and in capital letters? |
| **Unnamed questions** | Do you see names such as `Q17` or `Q23_1`? Those questions were never named. Fix the names in Qualtrics and export again. Do not rename them by hand in Excel. |
| **Minimum and maximum** | For each scale, do the minimum and maximum match the response options (for example, 1 to 5)? Are -99 and -50 gone? |
| **Team agreement** | Did every teammate import the same file, with the same date in its name? |
