---
title: "Visualize a Relationship"
subtitle: "How to build scatterplots to show how two scores go together"
author:
  - Andrew Silhavy
  - Shane McCarty, PhD
abstract: |
  This chapter teaches researchers to visualize relationship research questions, which examine how two continuous scores go together. Using real insurance data, researchers build a scatterplot one layer at a time: data, mapping, points, a line of best fit, grouping with color or facets, labels, a theme, and axis scales. In The Lab, researchers make a scatterplot with the lab dataset and then learn what to do when survey scores stack on top of each other: jitter the points and say so in the figure caption. In Your Turn, researchers use a copy-ready template and a checklist to build the relationship plot for their own research question.
keywords:
  - ggplot2
  - scatterplot
  - line of best fit
  - jitter
  - relationship
date: 09.20.2026
date-format: "MM.DD.YYYY"
format:
  html: 
    css: custom.css
editor: visual
execute:
  cache: false   # this chapter reads lab_prep.R and the lab dataset; a cache would keep OLD data after they change
  warning: false
  message: false
---

{{< include _ref-kickoff-box.qmd >}}

::: {.callout-note collapse="true"}
## Resources: `ggplot2`

{{< include _resources-ggplot2.qmd >}}
:::

## How This Chapter Works

You will go through this chapter three times, with three different datasets.

| Pass | Section | Dataset | What you do |
|------|-----------|-------------|--------------------------------------------|
| 1 | [The Play](#the-play) | `insurance` (example) | **Watch the play.** Read the code and the output, step by step. |
| 2 | [The Lab](#sec-relationship-lab) | lab dataset | **Run the play.** Make two scatterplots yourself during the R Lab, then check your own work. Labs are not graded. |
| 3 | [Your Turn](#sec-relationship-yourturn) | your team's dataset | **Call your own play.** After you finish all of the R Labs, come back and make the plot for your own research question. |

::: {.callout-tip}
## Go: Know your route first

This chapter is the Visualize step on [The Analysis Map](analysis-map.qmd) for a **relationship** question: two continuous scores. New to `ggplot2` layers? Read [Coding a Plot](layers.qmd) first.
:::

## 📋 The Play {#the-play}

### Example Data: `insurance`

The example dataset is from the "Medical Cost Personal Costs" database on Kaggle. For this chapter, we refer to it as `insurance`.

```{r}
#| label: import-data-insurance
insurance <- read.csv("data/insurance.csv")
```

```{r}
#| label: load-library-ggplot2
library(ggplot2)
```

### Relationship Research

For relationship research, the variables are 📏 **Continuous** because they range along a continuum, such as a scale for beliefs from 1 (strongly disagree) to 6 (strongly agree). To examine the relationship between two continous variables, use the steps below to build a custom scatterplot.

#### STEP-BY-STEP TO BUILD A SCATTERPLOT

#### [Step 1: Set up data]{.data-layer}

```{r}
ggplot(data = insurance)
```

#### [Step 2: Set up mapping]{.mapping-layer}

What is the relationship between BMI and insurance charges?

```{r}
ggplot(data = insurance,
       mapping = aes(
         x = bmi,
         y = charges))
```

::: callout-note
## 📊 Layer: Data + Mapping

Two layers are used:

- [**Data**]{style="color: #E69F00; font-weight: bold;"}: `insurance` dataset
- [**Mapping**]{style="color: #56B4E9; font-weight: bold;"}: `aes(x = bmi, y = charges)`
:::

#### [Step 3: Add geometry (points)]{.layers-layer}

Add points to create a scatterplot using `geom_point()`:

```{r}
ggplot(data = insurance,
       mapping = aes(
         x = bmi,
         y = charges)) + geom_point()
```

::: callout-note
## 🎨 Layer: Geometries

- [**Geometry**]{style="color: #009E73; font-weight: bold;"}: `geom_point()` displays data as points
:::

::: {.callout-warning}
## Caution: Don't start a line with `+`

Layers are joined with a `+` at the **end** of a line. If a line **starts** with `+`, R runs the plot without that layer and then gives you an error. Look at the code above: the `+` comes right before `geom_point()`, on the same line as the layer before it.
:::

#### [Step 4: Add a statistical layer (trend line)]{.statistics-layer}

Add a linear model line with `geom_smooth()`:

```{r}
ggplot(data = insurance,
       mapping = aes(
         x = bmi,
         y = charges)) + geom_point() +
  geom_smooth(method = "lm")
```

::: callout-note
## 📈 Layer: Statistics

- [**Statistics**]{style="color: #F0E442; font-weight: bold; text-shadow: 1px 1px 2px rgba(0,0,0,0.3);"}: `geom_smooth()` calculates and displays a trend line
:::

Notice how it looks like we have two different populations? Let's explore this further.

#### [Step 5. Incorporate grouping (color)]{.facets-layer}

If you have three variables (2 continouous, 1 categorical/grouping variable), then you can use facets. There are two main approaches to visualizing grouping variables:

##### Method 1: Using Color

Map the `smoker` variable to color:

```{r}
ggplot(data = insurance,
       mapping = aes(
         x = bmi,
         y = charges,
         color = smoker)) +
  geom_point() +
  geom_smooth(method = "lm")

```

::: {.callout-tip}
## Go: Put a variable inside `aes()`

`color = smoker` is inside `aes()` because `smoker` is a variable in the dataset. If you want every point to be one color, put the color **outside** `aes()`, like this: `geom_point(color = "#005a43")`.
:::

::: callout-note
## 🔵 Layer: Facet (extended)

- [**Facet**]{style="color: #0072B2; font-weight: bold;"}: Added `color = smoker` to group by smoking status
:::

##### Method 2: Using Facets

Create separate panels for each group with `facet_wrap()`:

```{r}
ggplot(data = insurance,
       mapping = aes(
         x = bmi,
         y = charges)) +
  geom_point() +
  geom_smooth(method = "lm") +
  facet_wrap(~smoker)
```

::: callout-note
## 📊 Layer: Facets

- [**Facets**]{style="color: #0072B2; font-weight: bold;"}: `facet_wrap(~smoker)` creates separate panels
:::

Both methods reveal that smoking status significantly affects the relationship between BMI and insurance charges!

#### [Step 6. Customize Appearance]{.theme-layer}

##### Adding Titles and Labels

Make your plot more informative with descriptive titles:

```{r}
ggplot(data = insurance,
       mapping = aes(
         x = bmi,
         y = charges)) +
  geom_point() +
  geom_smooth(method = "lm") +
  ggtitle("Insurance charges vs 
          BMI for smokers and non-smokers") +
  xlab("Body Mass Index (BMI)") +
  ylab("Insurance Charges")
```

##### Changing the Theme

Apply a professional theme with `theme_bw()`:

```{r}
#| label: ggplot with black and white theme 
#| fig-cap: "a scatterplot depicting a linear relationship between BMI and insurance charges"
#| fig-height: 6
#| fig-width: 8

ggplot(data = insurance, mapping = aes(x = bmi, y = charges, color = smoker)) +
  geom_point() +
  geom_smooth(method = "lm") +
  ggtitle("Insurance charges vs 
          BMI for smokers and non-smokers") +
  xlab("Body Mass Index (BMI)") +
  ylab("Insurance Charges") +
  theme_bw()
```

::: callout-note
## 🎨 Layer: Theme

- [**Theme**]{style="color: #CC79A7; font-weight: bold;"}: `theme_bw()` controls the overall visual appearance
:::

::: {.callout-important .required}
## Required for Team Poster: pick one theme for your team

Team members should pick **one theme** (from the [complete themes](https://ggplot2.tidyverse.org/reference/ggtheme.html)) to use with all plots in your individual reports and the team poster!
:::

##### [Adjusting Axis Scales]{.scales-layer}

Sometimes, you need to manually control the range of your axes for better visualization or to match other plots or to capture the entire range of the possible response options (e.g., 1-6, 1-10, 1 - 100).

Set specific limits with `xlim()` and `ylim()`.

```{r}
ggplot(data = insurance, 
       mapping = aes(x = bmi, y = charges, color = smoker)) +
  geom_point() +
  geom_smooth(method = "lm") +
  xlim(15, 55) +  # Set x-axis from 15 to 55
  ylim(0, 65000) +  # Set y-axis from 0 to 65000
  ggtitle("Insurance charges vs BMI for smokers and non-smokers") +
  xlab("Body Mass Index (BMI)") +
  ylab("Insurance Charges") +
  theme_bw()
```

::: callout-note
## 📏 Layer: Scales

- [**Scales**]{style="color: #F0E442; font-weight: bold; text-shadow: 1px 1px 2px rgba(0,0,0,0.3);"}: `xlim()` and `ylim()` control axis ranges
:::

::: {.callout-warning}
## Caution: Data outside the limits will be removed

Using `xlim()` and `ylim()` removes any data points outside the specified range. This can affect trend lines!
:::

::: {.callout-important .required}
## Required: x and y axes cover the full scale

If your response options range from 1 to 6. Use `ylim( )` to adjust the y-axis to range from 1 to 6 too.
:::

::: {.callout-tip}
## Go: Save your plot with `ggsave()`

Every plot in your *RD Report* and *Final Report* must be saved with `ggsave()`, so you can add it to your team poster. Save the plot to an object, then add two lines:

``` r
ggsave("plots/plot1_clear_name.png",
       plot = plot_name, width = 10, height = 8, dpi = 300)
```

The full explanation is in [Coding a Plot](layers.qmd#save).
:::

```{r}
#| label: lab-setup
#| include: false
source("lab_prep.R")   # creates mh_raw and mh_clean from the lab dataset
```

## 🏈 The Lab {#sec-relationship-lab}

You watched the play above with the `insurance` dataset (Pass 1). Now run two plays yourself with the **lab dataset**, a survey about what people believe about mental health.

:::::: {.ref .ref-tip}
::: ref-face
![](images/ref.svg){fig-alt="Ref the raccoon"}
:::

::: ref-say
**Ref's kickoff.** This lab is not graded, so I am here to help you check your own work. Try each play **before** you open my check or the solution. If your answer does not match, that is not a penalty. It is how you find out what to fix.
:::
::::::

| Lab play | What is new |
|----------------|------------------------------------------------|
| Screen Left | a scatterplot, just like the play above |
| Screen Right | points that stack on top of each other, and how to **jitter** them |

**Before you start**

1.  Open your RStudio Project and your lab `.qmd` file.
2.  Load the clean lab data by running this chunk. It runs the cleaning script you built in the earlier labs.

```{r}
#| label: lab-load
#| eval: false
source("lab_prep.R")   # creates mh_clean
nrow(mh_clean)         # how many people are in the clean dataset?
```

::: {.callout-tip collapse="true" icon="false"}
## ![](images/ref.svg){.ref-mini fig-alt="Ref the raccoon"} Ref's check: how many rows?

`nrow(mh_clean)` should be **`r nrow(mh_clean)`**. If you get `r nrow(mh_raw)`, you are looking at the raw data: the duplicate response and the people who failed the attention checks have not been removed yet.
:::

### Lab Play 1: Screen Left (a scatterplot)

**Research question:** Is age (`AGE`) related to the belief that substance use causes mental illness (`MIAQ_SUBSTANCE`, 1 to 7)?

**Your task:** Go back to [the scatterplot play](#the-play) and rebuild it for this question. Put `AGE` on the x axis and `MIAQ_SUBSTANCE` on the y axis, use `geom_point()`, add a line of best fit, label both axes in plain words, add a title, and use `theme_bw()`.

::: {.callout-tip collapse="true" icon="false"}
## ![](images/ref.svg){.ref-mini fig-alt="Ref the raccoon"} Ref's check: what should the plot look like?

```{r}
#| label: lab-check-play1
#| echo: false
lab_p1 <- mh_clean |> filter(!is.na(AGE), !is.na(MIAQ_SUBSTANCE))
lab_r1 <- cor(lab_p1$AGE, lab_p1$MIAQ_SUBSTANCE)
```

-   Your plot should include **`r nrow(lab_p1)`** people. R will warn you that it removed `r nrow(mh_clean) - nrow(lab_p1)` rows with missing values (two people typed an impossible age). That warning is expected.
-   The line of best fit should slope gently **`r ifelse(lab_r1 < 0, "downward", "upward")`**, and the dots should be spread widely around it. This is what a *weak* relationship looks like (r = `r sprintf("%.2f", lab_r1)`). Most relationships in survey research look more like this than like a tight line.
-   You can see almost every person as a separate dot, because `AGE` can take many different values.
:::

::: {.callout-note .gray collapse="true"}
## 👀 See the solution

```{r}
#| label: plot-lab-age-substance
#| fig-cap: "Figure 1. Age and the belief that substance use causes mental illness in the lab dataset. The line is a linear line of best fit."
#| fig-alt: "A scatterplot with age on the x axis and the substance use belief score on the y axis. The points are widely spread and the line of best fit slopes gently upward."
#| fig-height: 5
#| fig-width: 7

plot_age_substance <- ggplot(
  data = mh_clean,
  mapping = aes(
    x = AGE,
    y = MIAQ_SUBSTANCE)) +
  geom_point() +
  geom_smooth(method = "lm") +
  ylim(1, 7) +
  ggtitle("Age and the Belief that Substance Use Causes Mental Illness") +
  xlab("Age (years)") +
  ylab("Substance use as a cause (1 = not important, 7 = very important)") +
  theme_bw()

print(plot_age_substance)
```
:::

### Lab Play 2: Screen Right (a scatterplot with stacked points)

**Research question:** Is mental health stigma (`STIGMA`, 1 to 5) related to mental health self-efficacy (`EFFICACY`, 1 to 6)?

**Your task, part 1:** Make the same kind of scatterplot as Lab Play 1, with `STIGMA` on the x axis and `EFFICACY` on the y axis. Use `geom_point()`. Then count the dots. Does it look like `r nrow(mh_clean)` people?

```{r}
#| label: plot-lab-stacked
#| echo: false
#| fig-cap: "With geom_point(), people who have the same two scores are drawn on top of each other."
#| fig-alt: "A scatterplot of stigma and self-efficacy where the points sit in a grid pattern and many points hide other points."
#| fig-height: 4
#| fig-width: 6
ggplot(data = mh_clean, mapping = aes(x = STIGMA, y = EFFICACY)) +
  geom_point() +
  ylim(1, 6) +
  theme_bw()
```

::: {.callout-warning icon="false"}
## ![](images/ref-whistle.svg){.ref-mini fig-alt="Ref the raccoon blowing his whistle"} Foul: stacked points

`EFFICACY` is the average of three survey items on a 6-point scale, so it can only take a few values (1, 1.33, 1.67, 2, and so on). `STIGMA` is the same kind of variable. When two people have the same two scores, `geom_point()` draws one dot **on top of** the other, and your plot hides most of your sample. This happens with almost every survey scale, so check for it every time.
:::

**Your task, part 2:** Fix it. Replace `geom_point()` with `geom_jitter()`. Jitter nudges each dot a tiny random distance so that stacked dots separate. Adding `alpha = 0.5` makes the dots see-through, so darker areas show where more people are.

```{r}
#| label: lab-jitter-hint
#| eval: false
  geom_jitter(width = 0.08, height = 0.08, alpha = 0.5) +
```

::: {.callout-warning}
## Caution: Jitter a little, and say that you did

Jitter only moves the dots in the *picture*. It does not change your data or your line of best fit. Keep `width` and `height` small (less than half the distance between two possible scores), and tell your reader in the figure caption that the points are jittered.
:::

::: {.callout-tip collapse="true" icon="false"}
## ![](images/ref.svg){.ref-mini fig-alt="Ref the raccoon"} Ref's check: how many dots can you see?

```{r}
#| label: lab-check-play2
#| echo: false
lab_p2 <- mh_clean |> filter(!is.na(STIGMA), !is.na(EFFICACY))
lab_r2 <- cor(lab_p2$STIGMA, lab_p2$EFFICACY)
lab_spots <- lab_p2 |> count(STIGMA, EFFICACY)
```

-   There are **`r nrow(lab_p2)`** people in this plot, but with `geom_point()` you can only see **`r nrow(lab_spots)`** dots. The biggest stack has `r max(lab_spots$n)` people on a single dot.
-   After you switch to `geom_jitter()`, you should see small clouds of dots where the single dots used to be. Your dots will not be in exactly the same places as the solution, because jitter is random. That is fine.
-   The line of best fit should slope **`r ifelse(lab_r2 < 0, "downward", "upward")`**: people who report more stigma report `r ifelse(lab_r2 < 0, "less", "more")` self-efficacy (r = `r sprintf("%.2f", lab_r2)`). The line is the same with or without jitter.
:::

::: {.callout-note .gray collapse="true"}
## 👀 See the solution

```{r}
#| label: plot-lab-stigma-efficacy
#| fig-cap: "Figure 2. Mental health stigma and mental health self-efficacy in the lab dataset. Points are jittered to show overlapping responses; the line is a linear line of best fit."
#| fig-alt: "A scatterplot with stigma on the x axis and self-efficacy on the y axis. Jittered, see-through points form small clouds. The line of best fit slopes downward."
#| fig-height: 5
#| fig-width: 7

plot_stigma_efficacy <- ggplot(
  data = mh_clean,
  mapping = aes(
    x = STIGMA,
    y = EFFICACY)) +
  geom_jitter(width = 0.08, height = 0.08, alpha = 0.5) +
  geom_smooth(method = "lm") +
  xlim(1, 5) +
  ylim(1, 6) +
  ggtitle("More Stigma, Less Mental Health Self-Efficacy") +
  xlab("Mental health stigma (1 = low, 5 = high)") +
  ylab("Mental health self-efficacy (1 = low, 6 = high)") +
  theme_bw()

print(plot_stigma_efficacy)
```
:::


::: {.callout-warning collapse="true" icon="false"}
## ![](images/ref-whistle.svg){.ref-mini fig-alt="Ref the raccoon blowing his whistle"} Ref's call: fouls to check if your plot does not match

-   **Your scatterplot looks like a neat grid of dots** → your points are stacked. Use `geom_jitter()` (Lab Play 2).
-   **Your jittered plot looks like a shapeless cloud** → your jitter is too big. Make `width` and `height` smaller.
-   **`object 'EFFICACY' not found`** → `EFFICACY` and `STIGMA` are composites created in `lab_prep.R`. Run `source("lab_prep.R")` first.
-   **An error that mentions `+`** → each ggplot line must *end* with `+`. A line that *starts* with `+` breaks the plot.
:::

:::::: {.ref .ref-tip}
::: ref-face
![](images/ref.svg){fig-alt="Ref the raccoon"}
:::

::: ref-say
**Ref's final whistle.** That is the end of the lab. Save your `.qmd`, render it, and back up your `.qmd` and `.R` files to the **R Labs** folder in your ELN before you quit RStudio.
:::
::::::

## 🏆 Your Turn {#sec-relationship-yourturn}

::: {.callout-note .gray icon="false"}
## There is no Ref. It's game time, your turn!

Nobody has analyzed your team's data before, so there is no answer to check against. Copy the play that matches your research question, change every line marked `# REPLACE`, and use the checklist at the end of this section to check your own plot. It is the same checklist your instructor uses.
:::

Once you finish all of the R Labs, come back here with **your team's dataset** and make the plot for your relationship research question.

### The relationship play

```{r}
#| label: plot-REPLACE             # REPLACE with a short name, e.g., plot-sleep-stress
#| fig-cap: "Figure 1. REPLACE with a caption that can stand alone."
#| fig-alt: "REPLACE with a description of what the plot shows."
#| eval: false

plot_REPLACE <- ggplot(                     # REPLACE the plot name
  data = yourdata,                          # REPLACE with your clean dataset
  mapping = aes(
    x = VARIABLE1,                          # REPLACE with your predictor
    y = VARIABLE2)) +                       # REPLACE with your outcome
  geom_point() +                            # points stacked? use geom_jitter(width = 0.08, height = 0.08, alpha = 0.5)
  geom_smooth(method = "lm") +
  ggtitle("REPLACE with your title") +
  xlab("REPLACE with a plain-words label and the scale range") +
  ylab("REPLACE with a plain-words label and the scale range") +
  theme_bw()                                # use the same theme in every plot

print(plot_REPLACE)
ggsave("plots/REPLACE.png",                 # REPLACE with a clear file name
       plot = plot_REPLACE,
       width = 10, height = 8, dpi = 300)
```


### Checklist for your RD Report and Final Report

::: {.callout-note .gray collapse="true"}
## Open the plot checklist

{{< include _checklist-plots.qmd >}}
:::

{{< include _ref-whistle-box.qmd >}}
