---
title: "Import Data Once"
subtitle: "How to use readr, readxl, and haven"
authors: 
  - Zihan Hei
  - Shane McCarty
abstract: |
  This chapter supports researchers in importing data from various sources and file formats commonly used in public health research. Researchers re-read the import line they first ran in Coding a Plot, learn its three parts (object, function, file path), import the same data as a .csv and as an Excel file, and inspect the result with head(), names(), nrow(), and ncol(). The chapter demonstrates essential data inspection techniques including viewing data structure, examining headers, and understanding data summaries. Researchers also learn to handle missing data by recoding special values (-99, -50) as NA during import. By the end of this chapter, researchers will be equipped to import data from multiple sources and prepare it for analysis in their public health research projects.
keywords:
  - data import
  - csv
  - excel
  - SPSS
  - missing data
date: 10.20.2025
date-format: "MM.DD.YYYY"
format:
  live-html:
    css: custom.css
engine: knitr
editor: visual
execute:
  cache: false   # reads the lab dataset; also, an old cache listed packages (NHANES) that are not installed on every computer
---

{{< include _ref-kickoff-box.qmd >}}

{{< include ./_extensions/r-wasm/live/_knitr.qmd >}}

::: {.callout-note collapse="true"}
## Resources: import

- [R4DS Book Chapter 4](https://r4ds.hadley.nz/data-import.html)
- [R4DS Book Chapter 20-24](https://r4ds.hadley.nz/import.html)
- [Posit Cheatsheet](https://rstudio.github.io/cheatsheets/html/data-import.html?_gl=1*2grk0d*_ga*ODQ0NjUzMTM1LjE3NjA0MDU5Nzk.*_ga_2C0WZ1JHG0*czE3NjA0OTMzNDgkbzMkZzAkdDE3NjA0OTMzNDgkajYwJGwwJGgw)
- [PDF Cheatsheet](https://rstudio.github.io/cheatsheets/data-import.pdf)
:::


## Introduction to Data Import

The first step of the data science workflow is to **import your data**: to bring a file from your project folder into R's memory, as an object you can use. Public health researchers import data from many sources: survey platforms (Qualtrics, REDCap), spreadsheets (Excel, Google Sheets), other statistical software (SPSS, SAS, Stata), and R packages. In FRI Public Health, your team's data comes from Qualtrics as an Excel file ([Export Survey Data](export.qmd)).

You have already imported data once. In [Coding a Plot](layers.qmd), the first line of code was:

```r
insurance <- read.csv("data/insurance.csv")
```

You watched it run without an explanation. Now you will read it, because every import you ever write has the same three parts.

## 📋 The Play {#the-play}

### Play 1: Read the import line

```{r}
#| label: play-import-csv
insurance <- read.csv("data/insurance.csv")
```

| Part | What it is | What it does |
|----|----|----|
| `insurance` | the **object** name you choose | the labeled box that will hold the data ([Coding a Plot](layers.qmd)) |
| `<-` | the **assignment arrow** ([Coding a Plot](layers.qmd)) | "put what is on the right into the box on the left" |
| `read.csv( )` | the **function** | reads a `.csv` file (comma-separated values, a plain-text spreadsheet) |
| `"data/insurance.csv"` | the function's **input**: a file path in quotes | the folder `data` inside your project, then the file name, then its extension `.csv` |

Read it from right to left: open the file `insurance.csv` in the folder `data`, read it as a table, and call that table `insurance`. R looks for the `data` folder starting from your project folder, which is why the project matters ([R Software](RStudio.qmd)). Nothing appears on the screen when this line runs. Look at the **environment tab** instead: `insurance` is there, with its number of rows (obs.) and columns (variables).

::: callout-important
## Important: make a `data` folder in your project, once, and put every data file in it

Every import line in this playbook starts with `"data/…"`. That path only works if there is a folder named **`data`** (lowercase, no space) inside your `Lastname_FRI` project folder, and the file is inside it. Neither exists until you make them:

1.  In RStudio, with your project open, go to the **Files** tab (bottom right) and click **New Folder**. Name it `data` and click OK. You do this **once**; the folder stays in your project.
2.  Put the data file inside that folder, not next to it. Do this in Finder or File Explorer: open the project folder, open `data`, and drag the file in ([R Software](RStudio.qmd) shows the steps).
3.  Check the **Files** tab: click `data` and you should see the file. Click the refresh icon at the top right of the tab if it does not appear.

**If the code will not run:** `path does not exist` means R could not find the file where the code says it is. Nine times out of ten the file is still in Downloads, or it is in the project folder but not inside `data`, or the folder is called `Data` or `data files`. Fix the folder, not the code. Keep `lab_prep.R` and your `.qmd` files in the project folder itself, next to `Lastname_FRI.Rproj`; only data goes in `data`.
:::

### Play 2: Look at what you imported

These are the same checks you ran in [Coding a Plot](layers.qmd), and you will run them on every dataset for the rest of the course. `head()` shows the first six rows. `names()` lists the variables. `nrow()` and `ncol()` count the people and the variables.

```{r}
#| label: play-head-csv
head(insurance)
```

```{r}
#| label: play-names-csv
names(insurance)
nrow(insurance)   # how many people?
ncol(insurance)   # how many variables?
```

`head()` is also how you catch the two most common import problems at once. Every row should be one person (one respondent, with their own `ResponseId`), so a first row that is question text instead of a person means the file still has two header rows ([Export Survey Data](export.qmd)). And every column of numbers should be numbers, so values in quotes, or `<chr>` under a column name instead of `<dbl>`, mean a numeric column came in as text.

**Try it.** Run `head(insurance)`, then change it to `head(insurance, 10)` and to `tail(insurance)`. The second input to `head()` is the number of rows; `tail()` shows the last rows instead of the first.

```{webr-r}
#| context: setup
download.file("https://raw.githubusercontent.com/shanemccarty/FRIplaybook/main/data/insurance.csv", "insurance.csv")
insurance <- read.csv("insurance.csv")
```

```{webr-r}
head(insurance)
```

### Play 3: Same play, Excel file

Your team's data is an Excel file (`.xlsx`), not a `.csv`. The import line has the **same three parts**. Only the function changes, because the file type changed, and that function comes from the `readxl` package, so `library(readxl)` has to run first ([R Packages](library.qmd)).

```{r}
#| label: play-import-xlsx
#| message: false
library(readxl)
insurance_xl <- read_excel("data/insurance.xlsx")
head(insurance_xl)
```

Compare the two `head()` outputs. Same people, same variables, same numbers. The file type does not change the data; it only changes which function reads it. (`read_excel()` prints the output as a *tibble*, which shows the number of rows and the type of each column, such as `<dbl>` for numbers and `<chr>` for text, under the column names. That extra line is useful, not a problem.)

For your team's data, the line looks like this (Team 1, exported on the October 9 lecture; the object is named `alldata` because it holds every response, and later chapters filter it down):

```r
alldata <- read_excel("data/10.09.2026.data.team1.clean.xlsx")
```

An Excel file can hold several sheets. `read_excel()` reads the **first** sheet unless you say otherwise, and Excel names that sheet `Sheet1` by default. Your `.clean` file has one sheet, and the lab file keeps its data on the first sheet, so in both cases the file name is enough. If you ever need a different sheet (the lab file also has a `codebook` sheet), add a second input, `sheet =`:

```r
mh_codebook <- read_excel("data/ANTH306_LayConceptionsMH_SYNTHETIC.xlsx", sheet = "codebook")
```

::: {.callout-important .required}
## Required: import from the `.clean` file, with the file name only

Your report imports your team's `.clean` file (one header row, [Export Survey Data](export.qmd)) from the `data` folder of your project, using the path `"data/…"` and nothing else: no `setwd()`, no `/Users/yourname/…`. That is what makes the same code run on your teammates' computers and your peer mentor and/or Dr. Shane's.
:::

::: callout-note
## Resources: other file types

The pattern holds for every file type: only the function changes. `read_sav()` from the `haven` package reads SPSS `.sav` files, `read_csv()` from `readr` (part of the tidyverse) is a faster `read.csv()`, and `read_dta()` reads Stata files. See the [Posit data import cheatsheet](https://rstudio.github.io/cheatsheets/html/data-import.html).
:::

```{r}
#| label: lab-import-setup
#| include: false
#| cache: false
library(readxl)
mh_raw_check <- read_excel("data/ANTH306_LayConceptionsMH_SYNTHETIC.xlsx")
mh_codebook_check <- read_excel("data/ANTH306_LayConceptionsMH_SYNTHETIC.xlsx", sheet = "codebook")
```

## 🏈 The Lab {#sec-import-lab}

You watched the play above with example datasets. Now import the **lab dataset** yourself: a survey about what people believe about mental health. Everyone in the course uses the same file, so you can check your answers.

:::::: {.ref .ref-tip}
::: ref-face
![](images/ref.svg){fig-alt="Ref the raccoon"}
:::

::: ref-say
**Ref's kickoff.** This lab is not graded, so I am here to help you check your own work. Try each step **before** you open my check. If your answer does not match, that is not a penalty. It is how you find out what to fix.
:::
::::::

**Before you start**

1.  Download the lab data file, `ANTH306_LayConceptionsMH_SYNTHETIC.xlsx`, from the class Google Drive. Put it in the **data** folder inside your `Lastname_FRI` project folder (the Important box above shows how to make that folder if you have not yet).
2.  From the same class Google Drive folder, download the short R script `lab_prep.R`. Put it in your `Lastname_FRI` project folder itself, next to `Lastname_FRI.Rproj` (not inside **data**). You will not use it in this chapter; from [Transforming Your Data](dplyr.qmd) on, every Lab starts by running it.
3.  Open your RStudio Project and your lab `.qmd` file, and run your `library()` chunk. You need `readxl`.

### Lab Play 1: Import the data

The file has more than one sheet. The survey answers are on the first sheet, `Sheet1`, which is the one `read_excel()` reads by default. Import it and name the object `mh_raw`. Then find out how many rows (people) and columns (variables) it has.

::: {.callout-tip collapse="true" icon="false"}
## ![](images/ref.svg){.ref-mini fig-alt="Ref the raccoon"} Ref's check: how big is the dataset?

`mh_raw` should have **`r nrow(mh_raw_check)` rows** (people) and **`r ncol(mh_raw_check)` columns** (variables). Look in your **environment tab**: it shows the same two numbers next to `mh_raw`, as `r nrow(mh_raw_check)` obs. of `r ncol(mh_raw_check)` variables.
:::

::: {.callout-note .gray collapse="true"}
## 👀 See the solution

```{r}
#| label: lab-import-solution
#| eval: false
library(readxl)

mh_raw <- read_excel("data/ANTH306_LayConceptionsMH_SYNTHETIC.xlsx")

nrow(mh_raw)   # how many people?
ncol(mh_raw)   # how many variables?
```
:::

### Lab Play 2: View the variables

Look at the column names with `names()`, look at the first rows with `head()`, and get a summary of one variable, `POLITICALBELIEFS`, with `summary(mh_raw$POLITICALBELIEFS)`.

::: {.callout-tip collapse="true" icon="false"}
## ![](images/ref.svg){.ref-mini fig-alt="Ref the raccoon"} Ref's check: what should you see?

-   The first five column names are: `r paste0("\x60", names(mh_raw_check)[1:5], "\x60", collapse = ", ")`. The last one is `r paste0("\x60", tail(names(mh_raw_check), 1), "\x60")`.
-   `POLITICALBELIEFS` has a minimum of `r min(mh_raw_check$POLITICALBELIEFS, na.rm = TRUE)` and a maximum of `r max(mh_raw_check$POLITICALBELIEFS, na.rm = TRUE)`.
-   A minimum of -99 is not a real answer. The answers to this question run from 1 to 6, and -99 is the code for "Prefer not to say". You will turn those codes into missing values (`NA`) in Lab Play 4.
:::

### Lab Play 3: Do the variable names match the survey?

This is the check that matters most. The column names in R come straight from the variable names that were typed into Qualtrics ([Name and Recode Variables in Qualtrics](qualtrics.qmd)). If a name is misspelled, or a question was never named, you will see it here first, and it is far cheaper to fix now than after you have written code around the wrong name.

For the lab dataset, the survey's variable names are listed on the sheet called `codebook`, one row per variable. Import that sheet as `mh_codebook`, then compare its `Variable` column with `names(mh_raw)`. Four things to look for:

1.  **Names that are not names.** Anything like `Q17`, `Q23_1`, or `QID9` is a question that was never named in Qualtrics. It has to be named there and exported again, not renamed by hand in Excel.
2.  **Spelling and capitals.** `STIGMA_PUB3` and `Stigma_Pub3` are two different variables to R. Watch for a stray space, a lowercase letter, or a typo (`STIMGA`).
3.  **Something missing.** A variable the codebook lists that is not in the data usually means a question was deleted or was not exported (a *Download all fields* problem, [Export Survey Data](export.qmd)).
4.  **Something extra.** Columns in the data that the codebook does not list. Some are normal: Qualtrics adds housekeeping columns (`StartDate`, `Duration_sec`, `Progress`), and a select-all question exports as one column per box (`RACIALIZED_1`, `RACIALIZED_2`, …) that a codebook often describes in a single row. Anything else extra needs an explanation.

Reading 134 names by eye is how you miss things, so let R do the comparison. `setdiff(A, B)` returns the items that are in `A` but **not** in `B`, so:

-   `setdiff(names(mh_raw), mh_codebook$Variable)` lists columns that are **in the data but not in the codebook** (checks 1, 2, and 4 above).
-   `setdiff(mh_codebook$Variable, names(mh_raw))` lists variables that are **in the codebook but not in the data** (check 3, and the other half of check 2: a misspelled name shows up in both lists, once under each spelling).

If both lists come back empty, the names match exactly. Run both on the lab data.

::: {.callout-tip collapse="true" icon="false"}
## ![](images/ref.svg){.ref-mini fig-alt="Ref the raccoon"} Ref's check: do they match?

```{r}
#| label: lab-import-check-names
#| echo: false
#| cache: false
cb_names <- mh_codebook_check$Variable
missing_from_data <- setdiff(cb_names, names(mh_raw_check))
```

-   `setdiff(names(mh_raw), mh_codebook$Variable)` (in the data, not in the codebook) returns **`r length(setdiff(names(mh_raw_check), cb_names))` names**, and all of them start with `r paste0("\x60", unique(sub("_[0-9]+$", "", setdiff(names(mh_raw_check), cb_names))), "\x60", collapse = ", ")`. That is check 4, the harmless kind: a select-all question that exports one column per box, which the codebook describes in one row. No `Q17`, no misspellings.
-   `setdiff(mh_codebook$Variable, names(mh_raw))` (in the codebook, not in the data) returns **`r length(setdiff(cb_names, names(mh_raw_check)))` names**: `r paste0("\x60", setdiff(cb_names, names(mh_raw_check)), "\x60", collapse = ", ")`. Those are not variables; they are the codebook's shorthand rows that describe a whole group of columns in one line (the MIAQ items, and the select-all religious identity boxes). The same harmless mismatch, seen from the other side.
-   So the lab data pass. That is on purpose: the lab file was named correctly so that you can learn the check on clean data. Your team's export is where `Q17` will show up, and this is the play that catches it.
:::

::: {.callout-note .gray collapse="true"}
## 👀 See the solution

```{r}
#| label: lab-import-solution-names
#| eval: false
names(mh_raw)                  # the VARIABLES in your dataset
head(mh_raw, 10)               # the first 10 people
summary(mh_raw$POLITICALBELIEFS)   # one variable: look at the minimum

mh_codebook <- read_excel("data/ANTH306_LayConceptionsMH_SYNTHETIC.xlsx",
                          sheet = "codebook")

# in the data but NOT in the codebook: unnamed questions, typos, extra columns
setdiff(names(mh_raw), mh_codebook$Variable)

# in the codebook but NOT in the data: missing or misspelled variables
setdiff(mh_codebook$Variable, names(mh_raw))
```
:::

### Lab Play 4: Turn the missing codes into `NA`

Lab Play 2 found a minimum of `-99` in `POLITICALBELIEFS`. In this survey, `-99` means "Prefer not to say" and `-50` means "Don't know" ([Name and Recode Variables in Qualtrics](qualtrics.qmd)). If they stay as numbers, R will average them into every score. Replace both codes with `NA`, R's word for missing, everywhere in the dataset, then run `summary(mh_raw$POLITICALBELIEFS)` again.

::: {.callout-tip collapse="true" icon="false"}
## ![](images/ref.svg){.ref-mini fig-alt="Ref the raccoon"} Ref's check: did the codes go away?

-   `summary(mh_raw$POLITICALBELIEFS)` now shows a minimum of **`r min(mh_raw_check$POLITICALBELIEFS[!mh_raw_check$POLITICALBELIEFS %in% c(-99, -50)], na.rm = TRUE)`** and a new entry, `NA's`, with the number of people who did not answer.
-   The number of rows has **not** changed. Replacing a value with `NA` keeps the person; it only blanks that one answer.
:::

::: {.callout-note .gray collapse="true"}
## 👀 See the solution

```{r}
#| label: lab-import-solution-na
#| eval: false
mh_raw[mh_raw == -99] <- NA    # prefer not to say -> missing
mh_raw[mh_raw == -50] <- NA    # don't know -> missing

summary(mh_raw$POLITICALBELIEFS)
```

`mh_raw == -99` finds every cell equal to -99, and `<- NA` writes `NA` into those cells. [Transforming Your Data](dplyr.qmd) shows the tidyverse way to do the same thing, `replace_with_na_all()`.
:::

::: {.callout-warning collapse="true" icon="false"}
## ![](images/ref-whistle.svg){.ref-mini fig-alt="Ref the raccoon blowing his whistle"} Ref's call: fouls to check if your numbers do not match

-   **`path does not exist`** → the file is not where the code says it is. Check that the file is inside a **data** folder in your project, that the name is spelled exactly the same, and that you opened your **project** (not only a file).
-   **You have 1 column, or the wrong columns** → you imported the wrong sheet. `read_excel()` reads the first sheet unless you add `sheet =`; check that your data are on the first sheet, or name the sheet you want.
-   **`could not find function "read_excel"`** → run `library(readxl)`.
-   **Your first row of data is question text** → the file still has two header rows. See [Export Survey Data](export.qmd) and delete row 2 in the `.clean` file. (The lab dataset has one header row, so you should not see this in the lab.)
:::

:::::: {.ref .ref-tip}
::: ref-face
![](images/ref.svg){fig-alt="Ref the raccoon"}
:::

::: ref-say
**Ref's final whistle.** That is the end of the lab. Save your `.qmd`, render it, and back up your `.qmd` and `.R` files to the **R Labs** folder in your ELN before you quit RStudio.
:::
::::::

## 🏆 Your Turn {#sec-import-yourturn}

::: {.callout-note .gray icon="false"}
## There is no Ref. It's game time, your turn!

Nobody has imported your team's data before, so there is no answer to check against. Use the checklist below.
:::

::: callout-warning
## Caution: meet with a Quant Peer Mentor before you analyze your team data

After data collection ends and you have downloaded your team dataset, meet with a Quant Peer Mentor during lab before you go further. The quant PM lab schedule is in Column E of the "Student Lab Hours (ANTH306-f2026)" Google Sheet.
:::

Once you finish all of the R Labs, come back here with **your team's `.clean` data file**. Name the object `alldata`: it holds every response, and later chapters filter it down (for example, to serious responses only) into smaller objects with their own names. The example below is for Team 1 exporting on the October 9 lecture; change the team number and the date to match your file.

```{r}
#| label: import-data-team
#| eval: false
library(readxl)

alldata <- read_excel("data/10.09.2026.data.team1.clean.xlsx")   # Team 1 example: use your team number and export date

alldata[alldata == -99] <- NA    # prefer not to say -> missing
alldata[alldata == -50] <- NA    # don't know -> missing

names(alldata)      # the VARIABLES
head(alldata, 10)   # the first 10 people
summary(alldata)    # minimum and maximum of every variable
```

| Criteria | Ask yourself |
|------------|--------------------------------------------------------|
| **One header row** | Is row 1 the variable names, and is row 2 a person (not question text)? |
| **Rows** | Does the number of rows match the number of responses in Qualtrics? |
| **Variable names** | Open your exported survey (`.docx`) next to `names(alldata)`. Does every column name match the variable name you chose in Qualtrics, spelled the same way and in capital letters? |
| **Unnamed questions** | Do you see names such as `Q17` or `Q23_1`? Those questions were never named. Fix the names in Qualtrics and export again. Do not rename them by hand in Excel. |
| **Minimum and maximum** | For each scale, do the minimum and maximum match the response options (for example, 1 to 5)? Are -99 and -50 gone? |
| **Team agreement** | Did every teammate import the same file, with the same date in its name? |

{{< include _ref-whistle-box.qmd >}}
