---
title: "Viridis Color Palettes"
author: "Your name"
date: "`r Sys.Date()`"
format: revealjs
editor: visual
---

```{r}
#| label: setup
#| include: false

# Set global knit options (optional)
knitr::opts_chunk$set(echo = FALSE)

# load packages
library(tidyverse)
library(viridis)
library(here)
library(maps)
library(sf)

# load data and functions
source(here("R/plot_temps.R"))
```

<!--Each second level header begins a new slide-->

## Viridis

The [viridis](https://sjmgarnier.github.io/viridis/) package provides eight eye-catching color palettes to use in data visualizations:

-   magma
-   inferno
-   plasma
-   viridis
-   cividis
-   rocket
-   mako
-   turbo

## Optimized for perception

According to the virids documentation, the palettes:

> are designed to improve graph readability for readers with common forms of color blindness and/or color vision deficiency. The color maps are also perceptually-uniform, both in regular form and also when converted to black-and-white for printing.

## Magma

```{r}
#| label: magma
plot_temps("magma", title = FALSE)
```

## Inferno

```{r}
#| label: inferno
plot_temps("inferno", title = FALSE)
```

## Plasma

```{r}
#| label: plasma
plot_temps("plasma", title = FALSE)
```

## Viridis

```{r}
#| label: viridis
plot_temps("viridis", title = FALSE)
```

## Cividis

```{r}
#| label: cividis
plot_temps("cividis", title = FALSE)
```

## Rocket

```{r}
#| label: rocket
plot_temps("rocket", title = FALSE)
```

## Mako

```{r}
#| label: mako
plot_temps("mako", title = FALSE)
```

## Turbo

```{r}
#| label: turbo
plot_temps("turbo", title = FALSE)
```

## Usage

To use a viridis palette in your plot, include `scale_*_viridis()` in your ggplot2 call:

```{r}
#| eval: false
#| echo: true

mpg |>  
  ggplot(aes(x = displ, y = cty, color = hwy)) +
    geom_point() +
    scale_color_viridis(option = "A")
```
