---
title: "Transforming Your Data"
subtitle: "How to use dplyr"
author: "Mara Estreich"
abstract: |
  This chapter introduces students to dplyr, a core package within the tidyverse designed
  for efficient data manipulation and transformation. Students learn essential data cleaning
  techniques including filtering observations, recoding variables, and creating new categorical
  and binary variables. Through practical examples using survey data, the chapter demonstrates
  how to prepare datasets for analysis by grouping response categories and creating dummy
  variables for regression modeling. By the end of this chapter, students will understand
  that clean, well-organized data is foundational for accurate analysis and will be equipped
  with core dplyr functions to transform raw survey data into analysis-ready datasets.
keywords:
  - data manipulation
  - dplyr
  - recoding variables
  - creating new variables
  - dummy coding
format: html
editor: visual
execute:
  warning: false
  message: false
---

{{< include _ref-kickoff-box.qmd >}}

::: {.callout-note collapse="true"}
## Resources: `dplyr`

-   [Documentation](https://dplyr.tidyverse.org/index.html)
-   [R4DS Book Chapter 3](https://r4ds.hadley.nz/data-transform.html)
-   [R4DS Ch. 12 - 19](https://r4ds.hadley.nz/transform.html)
-   [Datacamp](https://www.datacamp.com/courses/data-manipulation-with-dplyr)
-   [Interactive Cheatsheet](https://rstudio.github.io/cheatsheets/html/data-transformation.html)
-   [PDF Cheatsheet](https://rstudio.github.io/cheatsheets/data-transformation.pdf)
:::

```{r}
#| label: load library
library(haven)
library(naniar)
library(dplyr)
library(knitr)
```

::: callout-note
## Resources: what is a `.sav` file, and why `read_sav()`?

Example 1 uses a dataset from an earlier FRI team that was saved in **SPSS** format, `.sav`. It is the same kind of thing as your team's `.xlsx` file, a table with one row per person, in a different file type. The pattern from [Import Data Once](import-once.qmd) holds: the import line has the same three parts, and only the function changes with the file type. `read_excel()` (package `readxl`) reads `.xlsx`, `read.csv()` reads `.csv`, and `read_sav()` (package `haven`) reads `.sav`. `user_na = FALSE` tells `read_sav()` to bring SPSS's "user missing" codes in as ordinary numbers, so that the next chunk can turn `-99` and `-50` into `NA` the same way you do for an Excel file. You will not need `read_sav()` for your own data unless your team receives an SPSS file.
:::

```{r}
#| label: import-dataset
alldata <- read_sav("data/10.31.2024.oralhealth.sav", user_na = FALSE)
```

```{r}
#| label: replace-missing-data

## replace with NA
alldata <- alldata %>%
  replace_with_na_all(condition = ~.x %in% c(-99, -50))
```

## Introduction to dplyr

dplyr is part of the tidyverse and provides intuitive functions for manipulating data. Think of it as a toolkit for cleaning and preparing your data for analysis.

**Why use dplyr?**

-   Makes data cleaning code readable and logical

-   Uses the pipe operator (`%>%`) to chain operations together

-   Essential for preparing data before creating visualizations or running analyses

> **Remember**: You cannot create meaningful visualizations or run valid statistical analyses without clean data!

## Example 1: Recoding Education Categories into Only 3 Education Categories

When working with survey data, we often need to collapse detailed response categories into broader groups. This is useful for:

-   Simplifying analysis

-   Creating groups with adequate sample sizes

-   Making results easier to interpret

### Original Education Variable

Our survey asked about education level with 7 response options:

1.  Less than high school

2.  High school diploma or GED

3.  Some college, but no degree

4.  Associate's or technical degree

5.  Bachelor's degree

6.  Master's degree

7.  Doctoral or professional degree

### Recoding into Three Categories

Let's group these into three meaningful categories:

```{r}
#| label: mutate-education-to-EDUCATION_3CAT
  
# Create a new 3-category education variable
alldata <- alldata %>%
  mutate(
    EDUCATION_3CAT = factor(
      case_when(
        Education_Level %in% c(1, 2, 3, 4) ~ "Less than Bachelor's",
        Education_Level == 5 ~ "Bachelor's Degree",
        Education_Level %in% c(6, 7) ~ "Graduate Degree"
      ),
      levels = c("Less than Bachelor's", "Bachelor's Degree", "Graduate Degree")
    )
  )

# Check the recoding with a cross-tabulation
table(alldata$Education_Level, alldata$EDUCATION_3CAT, useNA = "ifany")
```

## Example 2: Cut a Score into Categories (the K10)

Sometimes a published scale comes with **cut-off scores**, and a reader wants to know how many people fall in each band, not only the average. The lab dataset's `DISTRESS1` to `DISTRESS10` are the ten items of the Kessler Psychological Distress Scale (K10; Kessler et al., 2002). The K10 is scored by **adding** the ten items, for a total from 10 to 50 (`K10_TOTAL` in `lab_prep.R`). Andrews and Slade (2001) proposed the cut-offs that most surveys use, including the Victorian Population Health Survey (2001):

| K10 total | Likelihood of a mental disorder |
|---|---|
| 10 to 19 | likely to be well |
| 20 to 24 | likely to have a mild disorder |
| 25 to 29 | likely to have a moderate disorder |
| 30 to 50 | likely to have a severe disorder |

These are screening bands, not diagnoses: a score says how likely it is that someone would meet criteria in an interview, not that they do.

```{r}
#| label: setup-lab-k10
#| message: false
source("lab_prep.R")   # creates mh_clean, with K10_TOTAL
```

### Four categories: `DISTRESS_4CAT`

`case_when()` reads its conditions from the top down and stops at the first one that is true, so each line only has to name its upper limit. `factor()` with `levels =` puts the four groups in order, so that tables and plots list them from Well to Severe instead of alphabetically.

```{r}
#| label: mutate-k10-to-distress-4cat
mh_clean <- mh_clean |>
  mutate(
    DISTRESS_4CAT = factor(
      case_when(
        K10_TOTAL <= 19 ~ "Well",
        K10_TOTAL <= 24 ~ "Mild",
        K10_TOTAL <= 29 ~ "Moderate",
        K10_TOTAL >= 30 ~ "Severe"
      ),
      levels = c("Well", "Mild", "Moderate", "Severe")
    )
  )

# ALWAYS check a transformation: is every score in the right band?
mh_clean |>
  group_by(DISTRESS_4CAT) |>
  summarise(n = n(), lowest = min(K10_TOTAL), highest = max(K10_TOTAL))

#source: Andrews & Slade (2001) cut-offs; R4DS Ch. 3 (Wickham et al., 2023) https://r4ds.hadley.nz/data-transform.html
#explanation: K10_TOTAL is cut into the four published bands; the summary table proves that each band's lowest and highest scores fall inside its limits
```

### Two categories: `DISTRESS_01CAT`

For a comparison of two groups ([Compare Groups](visualize-comparison.qmd)), or as a predictor in a regression ([Relate 3+ Variables](regression.qmd)), a **binary** version is often all you need. The K10 cut-off for "likely to have a disorder" is 20 or more, so the line falls between 19 and 20 (a score of 19.5 is impossible, since the total is a whole number). A binary variable is coded **0 and 1** and named for the 1: here, 1 means likely distressed.

```{r}
#| label: mutate-k10-to-distress-01cat
mh_clean <- mh_clean |>
  mutate(DISTRESS_01CAT = if_else(K10_TOTAL >= 20, 1, 0))

mh_clean |> count(DISTRESS_4CAT, DISTRESS_01CAT)    # Well -> 0; Mild, Moderate, Severe -> 1

#source: The Quantitative Playbook for R (McCarty, 2026)
#explanation: if_else() gives a 1 to everyone at or above the cut-off and a 0 to everyone below it; the count shows that the binary variable lines up with the four bands
```

## Example 3: Combine Select-All Boxes into One Variable (RACIALIZED)

The race/ethnicity question let people check more than one box, and Qualtrics exported **one column per box**: `RACIALIZED_1` (American Indian or Alaska Native), `RACIALIZED_2` (Asian), `RACIALIZED_3` (Black or African American), `RACIALIZED_4` (Hispanic or Latino), `RACIALIZED_5` (Middle Eastern or North African), `RACIALIZED_6` (Native Hawaiian or Pacific Islander), `RACIALIZED_7` (White), `RACIALIZED_8` (Other, with the write-in in `RACIALIZED_8_TEXT`), and `RACIALIZED_99` (Prefer not to say). A cell holds the option's code when the box was checked and is blank (`NA`) otherwise. Look at a few rows first:

```{r}
#| label: look-racialized-boxes
mh_clean |>
  select(RACIALIZED_1:RACIALIZED_99) |>
  head(8)
```

Nine columns cannot go into a table or a model as they are. The plan: **count the boxes** each person checked, then use `case_when()` to sort people into groups, putting anyone who checked two or more boxes into their own group instead of forcing them into one.

### Step 1: count the checked boxes

`!is.na()` turns each cell into `TRUE` (checked) or `FALSE` (blank), and `rowSums()` adds up the `TRUE`s across the eight identity columns. `RACIALIZED_99` is left out: declining to answer is not an identity.

```{r}
#| label: mutate-racialized-n
mh_clean <- mh_clean |>
  mutate(RACIALIZED_N = rowSums(!is.na(across(RACIALIZED_1:RACIALIZED_8))))

mh_clean |> count(RACIALIZED_N)
```

### Step 2: sort people into groups: `RACIALIZED_6CAT`

The **first** condition that is true wins, so "two or more" has to come first; otherwise a person who checked Asian and White would land in whichever single-box line came first. Groups with very few people (codes 1, 5, 6, and 8) are combined into "Another identity", because a group of three cannot support a comparison. `factor()` with `levels =` puts White first, which makes it the comparison group in every model later.

```{r}
#| label: mutate-racialized-6cat
mh_clean <- mh_clean |>
  mutate(
    RACIALIZED_6CAT = factor(
      case_when(
        RACIALIZED_N >= 2     ~ "Two or more",
        !is.na(RACIALIZED_7)  ~ "White",
        !is.na(RACIALIZED_2)  ~ "Asian",
        !is.na(RACIALIZED_3)  ~ "Black",
        !is.na(RACIALIZED_4)  ~ "Hispanic or Latino",
        RACIALIZED_N == 1     ~ "Another identity"      # codes 1, 5, 6, 8: too few to stand alone
      ),
      levels = c("White", "Asian", "Black", "Hispanic or Latino", "Another identity", "Two or more")
    )
  )

# ALWAYS check a transformation
mh_clean |> count(RACIALIZED_6CAT)

#source: The Quantitative Playbook for R (McCarty, 2026); R4DS Ch. 3 https://r4ds.hadley.nz/data-transform.html
#explanation: people who checked two or more boxes become one group; single-box people are sorted by which box; the four smallest groups are combined; people who only declined get NA
```

People who checked only "Prefer not to say" match none of the conditions, so they get `NA`, which is what you want: a decline is missing, not a group.

### Step 3: the binary version: `RACIALIZED_01CAT`

Many research questions compare people racialized as white with everyone else. Build the 0/1 version **from** `RACIALIZED_6CAT`, so the two variables can never disagree.

```{r}
#| label: mutate-racialized-01cat
mh_clean <- mh_clean |>
  mutate(RACIALIZED_01CAT = case_when(RACIALIZED_6CAT == "White" ~ 0,
                                      !is.na(RACIALIZED_6CAT)    ~ 1))

mh_clean |> count(RACIALIZED_6CAT, RACIALIZED_01CAT)

#source: The Quantitative Playbook for R (McCarty, 2026)
#explanation: 0 = racialized as white, 1 = racialized as a person of color (every other group, including Two or more); NA stays NA
```

::: callout-warning
## Caution: the name is RACIALIZED, and the binary is a teaching example

The variable is called `RACIALIZED`, not `RACE`, because the survey recorded how people are classified within a system of racial sorting, not a trait they carry (Fields & Fields, 2012; Zuberi & Bonilla-Silva, 2008; D'Ignazio & Klein, 2020; see [Data Equity](data-equity.qmd)). When you report a result, write "people racialized as Black scored …", not "Black people are …". And `RACIALIZED_01CAT` is a coarse split that hides real differences between the groups it lumps together. It is here so that you can learn how a binary predictor works. A real report should also show `RACIALIZED_6CAT`.
:::

::: callout-tip
## Go: Name a new categorical variable for the number of groups it has

In this playbook, a variable made by cutting a score or collapsing categories ends in `_#CAT`, where # is the number of groups: `DISTRESS_4CAT`, `EDUCATION_3CAT`, `HEALTH_3CAT`, `POL_3CAT`, `RACIALIZED_6CAT`. A **binary** variable ends in `_01CAT`. It is either coded 0 and 1, or a factor with two levels, and either way the name says what the 1 (the second level) means: `DISTRESS_01CAT` (1 = likely distressed), `RACIALIZED_01CAT` (1 = racialized as a person of color, 0 = racialized as white). The name tells a reader what kind of variable it is before they look at a single value.
:::

::: {.callout-important .required}
## Required: cite the cut-offs

If you cut a published scale into categories in your report, cite where the cut-offs come from, the way the K10 references below do. Cut-offs you chose yourself are okay if there is existing theory to inform the decision or for practical purposes, but you must report what you did transparently in your methods section and with your R code.
:::

## References

D'Ignazio, C., & Klein, L. F. (2020). *Data feminism*. MIT Press. <https://data-feminism.mitpress.mit.edu/>

Fields, K. E., & Fields, B. J. (2012). *Racecraft: The soul of inequality in American life*. Verso.

Zuberi, T., & Bonilla-Silva, E. (Eds.). (2008). *White logic, white methods: Racism and methodology*. Rowman & Littlefield.

Andrews, G., & Slade, T. (2001). Interpreting scores on the Kessler Psychological Distress Scale (K10). *Australian and New Zealand Journal of Public Health, 25*(6), 494–497. <https://doi.org/10.1111/j.1467-842X.2001.tb00310.x>

Kessler, R. C., Andrews, G., Colpe, L. J., Hiripi, E., Mroczek, D. K., Normand, S.-L. T., Walters, E. E., & Zaslavsky, A. M. (2002). Short screening scales to monitor population prevalences and trends in non-specific psychological distress. *Psychological Medicine, 32*(6), 959–976. <https://doi.org/10.1017/S0033291702006074>

Kessler, R. C., Barker, P. R., Colpe, L. J., Epstein, J. F., Gfroerer, J. C., Hiripi, E., Howes, M. J., Normand, S.-L. T., Manderscheid, R. W., Walters, E. E., & Zaslavsky, A. M. (2003). Screening for serious mental illness in the general population. *Archives of General Psychiatry, 60*(2), 184–189. <https://doi.org/10.1001/archpsyc.60.2.184>

Victorian Department of Human Services. (2001). *Victorian Population Health Survey 2001: Selected findings*. Melbourne: Department of Human Services.

{{< include _ref-whistle-box.qmd >}}
