---
title: "Pre/Post Data"
subtitle: "How to link two surveys from the same people and reshape the data"
author:
  - Erica Sava
  - Gavin Rualo
  - Shane McCarty, PhD
abstract: |
  This chapter is for research teams who survey the same people twice, such as before and after a workshop or on the first and last day of a study. Researchers learn to link each person's two responses with a participant code (PASSWORD), to check the link for duplicates and for codes that do not match, and to reshape the combined data between wide format (one row per person, used for pre/post tests) and long format (one row per response, used for pre/post plots) with tidyr.
keywords:
  - Team 3
  - Team 5
  - pre/post
  - pretest
  - posttest
  - left_join
  - PASSWORD
  - tidyr
  - wide format
  - long format
date: 09.20.2026
date-format: "MM.DD.YYYY"
format:
  html:
    css: custom.css
editor: visual
execute:
  warning: false
  message: false
  cache: false
---

{{< include _ref-kickoff-box.qmd >}}

::: {.callout-note collapse="true"}
## Resources: joins and `tidyr`

- [Mutating joins in dplyr](https://dplyr.tidyverse.org/reference/mutate-joins.html)
- [R4DS Book Chapter 19: Joins](https://r4ds.hadley.nz/joins.html)
- [R4DS Book Chapter 5: Data tidying](https://r4ds.hadley.nz/data-tidy.html)
- [Datacamp: Reshaping Data with tidyr](https://www.datacamp.com/courses/reshaping-data-with-tidyr)
- [tidyr Cheatsheet](https://rstudio.github.io/cheatsheets/html/tidyr.html)
- [Reshaping data between long and wide formats (Dai, 2020)](https://shanghai.hosting.nyu.edu/data/r/reshaping.html)
:::

## Who needs this chapter?

You need this chapter if your team surveyed **the same people twice** and you want to know whether their scores changed. If your team has one survey, skip it.

::: {.callout-note .gray collapse="true"}
## 📁 Team 3: Day 1 and Day 14

Team 3 surveys the same participants on **Day 1** and on **Day 14**. Day 1 is your pretest and Day 14 is your posttest. Each participant types the same `PASSWORD` on both days, and you will use it to link their two responses.
:::

::: {.callout-note .gray collapse="true"}
## 📁 Team 5: before and after the workshop

Team 5 surveys participants **before the workshop** and **after the workshop**, on the same day. The pre-workshop survey is your pretest and the post-workshop survey is your posttest. Each participant types the same `PASSWORD` in both surveys, and you will use it to link their two responses.
:::

## The idea: one code links two responses

A pre/post study gives you two responses from each person. To compare them, R has to know which pretest belongs with which posttest. Names would work, but they are not anonymous. So each participant makes up a code, `PASSWORD`, and types the same code in both surveys. The code is what links the two responses.

Think of a coat check. You hand over your coat and get a ticket. Later, the ticket is the only thing that connects you to your coat. `PASSWORD` is the ticket.

## Two shapes of data

The same pre/post data can be arranged in two ways.

### Wide format: one row per person

| `PASSWORD` | `WELLBEING_pre` | `WELLBEING_post` |
|------------|-----------------|------------------|
| abc123     | 3.2             | 4.1              |
| gobills7   | 2.8             | 3.0              |

: *Wide format. Each row is a PARTICIPANT. The pretest and the posttest are in separate columns.*

Use wide format to **test** whether scores changed (a paired t-test or a Wilcoxon signed-rank test), because the test needs each person's two scores side by side.

### Long format: one row per response

| `PASSWORD` | `TIME` | `WELLBEING` |
|------------|--------|-------------|
| abc123     | pre    | 3.2         |
| gobills7   | pre    | 2.8         |
| abc123     | post   | 4.1         |
| gobills7   | post   | 3.0         |

: *Long format. Each row is a SURVEY RESPONSE, so each participant appears twice.*

Use long format to **plot** pre/post scores, because `ggplot2` needs one column for time (the x axis) and one column for the score (the y axis). See [Visualize Pre/Post Data](violin.qmd).

You will usually need both. How you get them depends on how your data arrive, and there are two situations:

| Situation | What you start with | What you do |
|----|----|----|
| **1. One file, long format** | One survey given twice and exported once. Every response is a row, and a column (here `TIME`) says whether it is pre or post. | Clean `PASSWORD`, then `pivot_wider()` to get one row per person. |
| **2. Two files, wide format** | Two separate surveys (or two exports), one for pre and one for post, each with one row per person. | Clean `PASSWORD` in both, `left_join()` them into one wide file, then `pivot_longer()` for plots. |

The Play shows both. The Lab practices situation 1 with the lab dataset. Your Turn has a template for each, so use the one that matches your team's data.

## 📋 The Play

The examples use very small made-up datasets, so that you can see every row.

### Situation 1: one file in long format

One survey, given twice, exported once. Each person has two rows, and `TIME` says which is which. Notice the two problems hiding in it: `Blue42` typed the password with different capitals the second time, and `sunny5` never took the posttest.

```{r}
#| label: load-library-prepost
library(dplyr)
library(tidyr)
```

```{r}
#| label: example-long
long_raw <- tibble(
  PASSWORD  = c("abc123", "gobills7", "Blue42", "pizza9", "sunny5",
                "abc123", "gobills7", "blue42", "pizza9"),
  TIME      = c("pre", "pre", "pre", "pre", "pre",
                "post", "post", "post", "post"),
  WELLBEING = c(3.2, 2.8, 3.5, 4.0, 2.5,
                4.1, 3.0, 3.9, 4.2),
  STRESS    = c(3.9, 4.2, 3.1, 2.8, 4.4,
                3.0, 4.0, 2.7, 2.5))
long_raw
```

**Step 1: clean the password.** To R, `Blue42` and `blue42` are two different people. Make every password lowercase and remove stray spaces before you do anything else.

```{r}
#| label: long-clean-password
long_data <- long_raw |>
  mutate(PASSWORD = tolower(trimws(PASSWORD)),
         TIME = factor(TIME, levels = c("pre", "post")))   # pre first, then post
```

**Step 2: check for duplicates.** In long format a duplicate is the same `PASSWORD` at the same `TIME`. There should be none; if there are, decide with your team which response to keep and use `distinct(PASSWORD, TIME, .keep_all = TRUE)`.

```{r}
#| label: long-duplicates
long_data |>
  count(PASSWORD, TIME) |>
  filter(n > 1)
```

**Step 3: `pivot_wider()`: one row per person.** `names_from` is the column that becomes part of the new column names (`TIME`), `values_from` lists the scores, and `names_glue` writes the names as `WELLBEING_pre`, `WELLBEING_post`, and so on.

```{r}
#| label: long-to-wide
wide_data <- long_data |>
  pivot_wider(
    names_from  = TIME,
    values_from = c(WELLBEING, STRESS),
    names_glue  = "{.value}_{TIME}")

wide_data
```

**Step 4: check the reshape.** The wide file should have one row per person, and anyone who took only one survey shows `NA` in the other time's columns.

```{r}
#| label: long-to-wide-check
nrow(wide_data)                          # one row per person
wide_data |> filter(is.na(WELLBEING_post))   # who has no posttest?
```

`sunny5` has no posttest, so their `_post` columns are `NA`. Now you have both shapes: `long_data` for plots and `wide_data` for the paired test.

### Situation 2: two files in wide format

Two separate surveys, each with one row per person. Here you have to **join** them, and the join is where mistakes hide.


```{r}
#| label: example-data-prepost
pretest <- tibble(
  PASSWORD  = c("abc123", "gobills7", "Blue42", "pizza9", "sunny5"),
  AGE       = c(19, 20, 19, 21, 20),
  WELLBEING = c(3.2, 2.8, 3.5, 4.0, 2.5),
  STRESS    = c(3.9, 4.2, 3.1, 2.8, 4.4))

posttest <- tibble(
  PASSWORD  = c("abc123", "gobills7", "blue42 ", "pizza9", "pizza9"),
  WELLBEING = c(4.1, 3.0, 3.9, 4.2, 4.4),
  STRESS    = c(3.0, 4.0, 2.7, 2.5, 2.4))
```

Look closely at the posttest. `sunny5` did not take it. `Blue42` typed the password differently the second time (`blue42` with a space at the end). `pizza9` took it twice. All three happen in real studies. The steps are the same as in situation 1, with a join in the middle.

### Step 1: Clean the password

Same as before, but in **both** datasets, before you join them.

```{r}
#| label: clean-password
pretest <- pretest |>
  mutate(PASSWORD = tolower(trimws(PASSWORD)))

posttest <- posttest |>
  mutate(PASSWORD = tolower(trimws(PASSWORD)))
```

### Step 2: Look for duplicates

```{r}
#| label: find-duplicates
posttest |>
  count(PASSWORD) |>
  filter(n > 1)
```

`pizza9` is in the posttest twice. Decide with your team which response to keep, use the same rule for everyone, and report it in your methods. A common rule is to keep the **first** response.

```{r}
#| label: remove-duplicates
posttest <- posttest |>
  distinct(PASSWORD, .keep_all = TRUE)    # keeps the first row for each PASSWORD
```

### Step 3: Join the two datasets by `PASSWORD`

`left_join()` keeps every row of the first dataset (the pretest) and adds the matching columns from the second (the posttest). Columns that have the same name in both files get an ending, so you can tell them apart.

```{r}
#| label: join-prepost
wide_data2 <- pretest |>
  left_join(posttest, by = "PASSWORD", suffix = c("_pre", "_post"))

wide_data2
```

This is **wide format**: one row per person, the same shape that `pivot_wider()` produced in situation 1. `AGE` was only in the pretest, so it keeps its name.

### Step 4: Check the join

::: callout-warning
## Caution: Always check a join

A join never gives you an error when people fail to match. It quietly fills in `NA`. You have to look.
:::

```{r}
#| label: check-join
nrow(pretest)      # people who took the pretest
nrow(wide_data2)   # should be the SAME number

# people who took the pretest and have no matching posttest
anti_join(pretest, posttest, by = "PASSWORD")
```

```{r}
#| label: check-join-values
#| echo: false
n_pre <- nrow(pretest); n_match <- sum(!is.na(wide_data2$WELLBEING_post))
```

Both numbers are `r n_pre`, so no rows were added by mistake. One person, `sunny5`, has no posttest, and their `_post` columns are `NA`. In your methods, report all three numbers: `r n_pre` people took the pretest, `r n_match` were matched to a posttest, and `r n_pre - n_match` could not be matched.

### Step 5: Reshape to long format for plots

`pivot_longer()` is the reverse of `pivot_wider()`: it takes the pre and post columns and stacks them. The part of the column name before the `_` becomes the score, and the part after it becomes a new column called `TIME`.

```{r}
#| label: wide-to-long
long_data2 <- wide_data2 |>
  pivot_longer(
    cols = c(WELLBEING_pre, WELLBEING_post, STRESS_pre, STRESS_post),
    names_to = c(".value", "TIME"),
    names_sep = "_") |>
  mutate(TIME = factor(TIME, levels = c("pre", "post")))   # pre first, then post

long_data2
```

Each person now has two rows, and the data are ready for [Visualize Pre/Post Data](violin.qmd).

::: callout-tip
## Go: Keep both datasets

Keep `wide_data` for your pre/post test and `long_data` for your pre/post plot. They hold the same information in two shapes.
:::

::: {.callout-note .gray collapse="true"}
## A longer example: three scales at two times (oral health)

In this example, Erica converts oral health data from wide to long format to compare three kinds of self-efficacy (BSE, CSE, ISE) at Time 1 and Time 2. The same steps work for studies with more than two time points.

```{{r}}
library(tidyr)

# Create long format with separate columns for each SE type
longdata <- pivot_longer(oralhealthdata_prepost,
                        cols = c("BSE_T1", "BSE_T2", "CSE_T1", "CSE_T2", "ISE_T1", "ISE_T2"),
                        names_to = c("SE_type", "Time"),
                        names_pattern = "(.+)_(.+)",
                        values_to = "SE",
                        values_drop_na = TRUE)

# Convert T1/T2 to before/after
longdata$Time <- ifelse(longdata$Time == "T1", "before", "after")

# Pivot wider to get separate columns for each SE type
longdata <- pivot_wider(longdata,
                       names_from = SE_type,
                       values_from = SE)

# Keep only the desired columns
longdata <- longdata[, c("Time", "BSE", "CSE", "ISE")]

longdata$Time <- factor(longdata$Time, levels = c('before','after'))

longdata_long <- longdata %>%
  pivot_longer(cols = c(BSE, CSE, ISE),
               names_to = "SelfEfficacy",
               values_to = "Score")

# Create a new variable combining 'SelfEfficacy' and 'Time'
longdata_long$SelfEfficacyTime <- factor(
  paste(longdata_long$SelfEfficacy, longdata_long$Time),
  levels = c("BSE before", "BSE after", "CSE before", "CSE after", "ISE before", "ISE after")
)
```
:::

## 🏈 The Lab {#sec-prepost-lab}

:::::: {.ref .ref-tip}
::: ref-face
![](images/ref.svg){fig-alt="Ref the raccoon"}
:::

::: ref-say
**Ref's kickoff.** This lab practices **situation 1**: one file in long format. The lab file `ANTH306_LayConceptionsMH_SYNTHETIC_LONG.xlsx` is the lab survey given twice, six weeks apart, and exported once. Every response is a row, `TIME` says `pre` or `post`, and `PASSWORD` links a person's two rows. Try each step before you open my check.
:::
::::::

```{r}
#| label: lab-prepost-setup
#| include: false
library(readxl)
lab_long_raw <- read_excel("data/ANTH306_LayConceptionsMH_SYNTHETIC_LONG.xlsx")
lab_long <- lab_long_raw |>
  mutate(PASSWORD = tolower(trimws(PASSWORD)), TIME = factor(TIME, levels = c("pre", "post")))
lab_dups <- lab_long |> count(PASSWORD, TIME) |> filter(n > 1)
lab_long1 <- lab_long |> distinct(PASSWORD, TIME, .keep_all = TRUE)
lab_wide <- lab_long1 |>
  select(PASSWORD, TIME, HELPSEEK) |>
  pivot_wider(names_from = TIME, values_from = HELPSEEK, names_glue = "HELPSEEK_{TIME}")
lab_both <- sum(!is.na(lab_wide$HELPSEEK_pre) & !is.na(lab_wide$HELPSEEK_post))
lab_pre_only <- sum(is.na(lab_wide$HELPSEEK_post)); lab_post_only <- sum(is.na(lab_wide$HELPSEEK_pre))
```

**Before you start.** Get `ANTH306_LayConceptionsMH_SYNTHETIC_LONG.xlsx` from your instructor and put it in the `data` folder of your project. Open your lab `.qmd`, and run your load-library chunk (you need `readxl`, `dplyr`, and `tidyr`).

### Lab Play 1: Import and look at the shape

Import the file as `lab_long`, then count the rows at each `TIME`. Look at the first rows: how many times does a person appear?

::: {.callout-tip collapse="true" icon="false"}
## ![](images/ref.svg){.ref-mini fig-alt="Ref the raccoon"} Ref's check: how many rows?

`r nrow(lab_long_raw)` rows in total: **`r sum(lab_long_raw$TIME == "pre")` pre** and **`r sum(lab_long_raw$TIME == "post")` post**. Not everyone came back, so there are fewer post rows than pre rows. That is normal.
:::

### Lab Play 2: Clean the password and look for duplicates

Make `PASSWORD` lowercase with no stray spaces, make `TIME` a factor with `pre` first, then count `PASSWORD` by `TIME` and filter to `n > 1`.

::: {.callout-tip collapse="true" icon="false"}
## ![](images/ref.svg){.ref-mini fig-alt="Ref the raccoon"} Ref's check: any duplicates?

One: `r paste0("\x60", lab_dups$PASSWORD, "\x60")` appears twice at `r lab_dups$TIME`. The same person submitted the pretest twice. Keep the first with `distinct(PASSWORD, TIME, .keep_all = TRUE)`.
:::

### Lab Play 3: Reshape to wide

Keep `PASSWORD`, `TIME`, and `HELPSEEK`, then `pivot_wider()` so that each person has `HELPSEEK_pre` and `HELPSEEK_post`. Then count three things: people with both scores, people with only a pretest, and people with only a posttest.

::: {.callout-tip collapse="true" icon="false"}
## ![](images/ref.svg){.ref-mini fig-alt="Ref the raccoon"} Ref's check: what should you get?

-   `r nrow(lab_wide)` rows, one per `PASSWORD`.
-   **`r lab_both` people** have both a pre and a post score. **`r lab_pre_only`** have only a pretest (they did not come back). **`r lab_post_only`** have only a posttest.
-   Ten people with only a posttest is suspicious: everyone who took the posttest was supposed to have taken the pretest. Look at their passwords. Most are one character away from a pretest password (a typo). Real teams have to decide, and write down, whether to match those by hand. For the lab, leave them unmatched.
:::

::: {.callout-note .gray collapse="true"}
## 👀 See the solution

```{r}
#| label: lab-prepost-solution
#| eval: false
library(readxl)
library(dplyr)
library(tidyr)

lab_long <- read_excel("data/ANTH306_LayConceptionsMH_SYNTHETIC_LONG.xlsx")
count(lab_long, TIME)                                 # Lab Play 1
head(lab_long)

lab_long <- lab_long |>                               # Lab Play 2
  mutate(PASSWORD = tolower(trimws(PASSWORD)),
         TIME = factor(TIME, levels = c("pre", "post")))
lab_long |> count(PASSWORD, TIME) |> filter(n > 1)
lab_long <- lab_long |> distinct(PASSWORD, TIME, .keep_all = TRUE)

lab_wide <- lab_long |>                               # Lab Play 3
  select(PASSWORD, TIME, HELPSEEK) |>
  pivot_wider(names_from = TIME, values_from = HELPSEEK, names_glue = "HELPSEEK_{TIME}")
nrow(lab_wide)
sum(!is.na(lab_wide$HELPSEEK_pre) & !is.na(lab_wide$HELPSEEK_post))   # both
sum(is.na(lab_wide$HELPSEEK_post))                                     # pre only
sum(is.na(lab_wide$HELPSEEK_pre))                                      # post only
```
:::

:::::: {.ref .ref-tip}
::: ref-face
![](images/ref-whistle.svg){fig-alt="Ref the raccoon blowing his whistle"}
:::

::: ref-say
**Ref's final whistle.** Keep `lab_wide` and `lab_long`: [Visualize Pre/Post Data](violin.qmd) plots the long one, and the paired test (Compare 1 Group, Pre/Post, coming soon) uses the wide one. Save, render, back up to your ELN.
:::
::::::

## 🏆 Your Turn {#sec-prepost-yourturn}

::: {.callout-note .gray icon="false"}
## There is no Ref. It's game time, your turn!

Only some teams have pre/post data. Use the template that matches how your data arrive, change every line marked `# REPLACE`, and use the checklist. Name the object `alldata` for the file you import, the same as in every other chapter.
:::

### Template 1: one file in long format

```{r}
#| label: prepost-long-template
#| eval: false
library(readxl)
library(dplyr)
library(tidyr)

# 1. import the one .clean file
alldata <- read_excel("data/10.09.2026.data.team3.clean.xlsx")   # REPLACE with your file

# 2. clean the password and put the times in order
alldata <- alldata |>
  mutate(PASSWORD = tolower(trimws(PASSWORD)),
         TIME = factor(TIME, levels = c("pre", "post")))      # REPLACE with your time column and its values

# 3. duplicates: same person, same time (then decide which response to keep)
alldata |> count(PASSWORD, TIME) |> filter(n > 1)
alldata <- alldata |> distinct(PASSWORD, TIME, .keep_all = TRUE)

# 4. wide format for the paired test
wide_data <- alldata |>
  pivot_wider(names_from = TIME,
              values_from = c(REPLACE1, REPLACE2),           # REPLACE with your pre/post scores
              names_glue = "{.value}_{TIME}")

# 5. check: one row per person; who is missing a time?
nrow(wide_data)
wide_data |> filter(is.na(REPLACE1_post))                     # REPLACE

# 6. long format for plots is alldata itself (already one row per response)
long_data <- alldata
```

### Template 2: two files in wide format (Teams 3 and 5)

```{r}
#| label: prepost-REPLACE
#| eval: false
library(readxl)
library(dplyr)
library(tidyr)

# 1. import both .clean files
pretest  <- read_excel("data/10.09.2026.data.team3.day1.clean.xlsx")    # REPLACE with your pretest file (Team 3: Day 1)
posttest <- read_excel("data/10.09.2026.data.team3.day14.clean.xlsx")   # REPLACE with your posttest file (Team 3: Day 14)

# 2. clean the password in both
pretest  <- pretest  |> mutate(PASSWORD = tolower(trimws(PASSWORD)))
posttest <- posttest |> mutate(PASSWORD = tolower(trimws(PASSWORD)))

# 3. look for duplicates in both (then decide which response to keep)
pretest  |> count(PASSWORD) |> filter(n > 1)
posttest |> count(PASSWORD) |> filter(n > 1)

# 4. join by PASSWORD
wide_data <- pretest |>
  left_join(posttest, by = "PASSWORD", suffix = c("_pre", "_post"))

# 5. check the join
nrow(pretest)
nrow(wide_data)
anti_join(pretest, posttest, by = "PASSWORD")

# 6. long format for plots
long_data <- wide_data |>
  pivot_longer(
    cols = c(REPLACE_pre, REPLACE_post),          # REPLACE with your pre and post columns
    names_to = c(".value", "TIME"),
    names_sep = "_") |>
  mutate(TIME = factor(TIME, levels = c("pre", "post")))
```

| Criteria | Ask yourself |
|----|----|
| **Same code in both files** | Is there a `PASSWORD` column in both datasets, spelled the same way? |
| **Clean passwords** | Did you make the passwords lowercase and remove extra spaces in **both** datasets before joining? |
| **Duplicates** | Did you look for passwords that appear more than once? Did your team agree on one rule for which response to keep? |
| **Same number of rows** | Does `nrow(wide_data)` equal `nrow(pretest)`? |
| **Unmatched people** | Did you look at the `anti_join()` list? Are any of them near matches (a typo) that your team should discuss? |
| **Reported in methods** | Do you report how many people took the pretest, how many were matched, and how many were not? |
| **Column names** | If a variable name already contains a `_` (for example, `SLEEP_HOURS_pre`), did you check that `pivot_longer()` split it in the right place? |
| **Both shapes kept** | Do you have `wide_data` for your test and `long_data` for your plot? |

{{< include _ref-whistle-box.qmd >}}
